---
# Render this project via `targets::tar_make()` (the `site` target in
# _targets.R), not `quarto render index.qmd` directly — this file reads
# pipeline outputs with tar_read()/tar_load(), so a direct render can use
# stale or missing data if the pipeline hasn't been run.
title: "Manuscript Title"
author:
  - name: "Your Name"
    affiliation: "Your Institution"
date: today
abstract: |
  Write a short abstract of the manuscript here.
---

# Introduction {@sec-intro}

Write your introduction here, citing sources as needed [@example2024].

```{r}
#| label: setup
#| include: false
library(targets)
```

# Methods

We can add full-line equations with `\$\$` and reference them [@eq-poisson]. This uses... LatEX? Mathjax?

$$
p(x) = \frac{e^{-\lambda} \lambda^{x}}{x !}
$$ {#eq-poisson}

# Results

You can refer back to sections [@sec-intro].

This is a figure @fig-example.

```{r}
#| label: fig-example
#| fig-cap: "Example figure generated from the pipeline. In general, figures should always have captions"
data <- tar_read(data)
model <- tar_read(model)
plot(data$x, data$y)
abline(model)
```

This is a table @tbl-example. For some reason, they use different caption variables.

```{r}
#| label: tbl-example
#| tbl-cap: "Example figure generated from the pipeline. In general, figures should always have captions"
data <- tar_read(data)
model <- tar_read(model)
plot(data$x, data$y)
abline(model)
```


# References

::: {#refs}
:::
