---
# Render via `targets::tar_make()` (the `render_manuscript` target in
# _targets.R), not `quarto render` directly - this file reads pipeline
# outputs via tar_read()/tar_load(), so a direct render can use stale or
# missing data.
title: "Manuscript Title"
# set author and affiliations in _quarto.yml
date: today
abstract: |
  Write a short abstract of the manuscript here.
# Scoped here, not project-wide in _quarto.yml - see the comment there
# for the real CI-confirmed bug this avoids for every other notebook.
bibliography: references.bib
csl: styles/apa.csl # other styles at https://github.com/citation-style-language/styles
---

::: {.callout-info}
## The Whole Game

I like writing manuscripts with Quarto. I also hate writing manuscripts with Quarto. (ditto `targets` for analysis pipelines) There's more to go wrong, and Murphy's law, it will. Every fix to an issue will introduce two more problems. What's the solution? Don't try and figure out how to write a manuscript in Quarto while writing a manuscript. Then you have two problems, and they're competing for your attention.

This repo is my (mainly Claude Code, to be honest) attempt to front-load all the frustrations and unexpected interactions of working with Quarto and targets, and to get out something that I like but is still acceptable to my Word-using colleagues. The emphasis is on empiricism, that is, we take nothing for granted and run everything ourselves (every version) to make sure it works. 

I plan to use it as a combination template and testbed for my favourite R and other packages. This means I'm *still* trying to do two things at once, which may turn out to be a mistake. Quarto 2 will probably break all of this when it comes out, but until then, here we are. Please feel free to reuse and critique. I don't know if this'll really be useful to anyone else, but let me know if it is!
:::

# Introduction {#sec-intro}

Write your introduction here, citing sources as needed [@example2024]. I personally prefer to use [Quarto Wingman](https://github.com/ntluong95/quarto-wingman)'s citation functionality, which hooks into the existing Zotero API and is very fast. Here are some more citations: [@hetheringtonTwoThreeFourAtom1975; @hartiganDipTestUnimodality1985; @tongEyesWideShut2024]:

`@example2024` - appears as @example2024
`[@example2024]` - appears as [@example2024]
`[@example2024; @hetheringtonTwoThreeFourAtom1975]` - appears as [@example2024; @hetheringtonTwoThreeFourAtom1975]

Note that depending on how you're handling your citation keys in Zotero, they may turn up as `nameYear` or `nameTitleTitleTitleYear` or whatever custom choice you set your BibTex keys to. 

Citations added inline are generally also added to your default or document-specific references file (set in [_quarto.yml](_quarto.yml), in this repo, `references.bib`, or the document frontmatter). Do note that because they're indexed by citation key, if you switch format part-way through you may end up with duplicate entries. 

```{r}
#| label: setup
#| include: false
#| warning: false

library(targets)
library(here)
here::i_am("manuscript/manuscript.qmd")
source(here("R/functions.R"))
```

```{r}
#| label: get-render-environment
#| include: false

# detect_render_format() (R/functions.R) always returns real, non-NULL
# values - see its roxygen comment for why a hand-rolled version of this
# previously broke with "object 'is_html' not found" whenever
# QUARTO_EXECUTE_INFO was unset.
fmt <- detect_render_format()
is_html <- fmt$is_html
is_typst <- fmt$is_typst
is_docx <- fmt$is_docx
is_interactive <- fmt$is_interactive
ext <- fmt$ext
```

# Methods

We can add full-line equations with `\$\$` and reference them [@eq-poisson]. By default, they are rendered by MathJaX (what version? from where?).

$$
p(x) = \frac{e^{-\lambda} \lambda^{x}}{x !}
$$ {#eq-poisson}

# Results

You can refer back to sections [@sec-intro]. We had to put `{#sec-intro}` after "Introduction" in the Introduction header above to make this work.

This is a figure @fig-01-example.

{{< include figures/_fig-01-example.qmd >}}

This is a table @tbl-01-example. For some reason, they use different caption variables.

{{< include tables/_tbl-01-example.qmd >}}

::: {.callout-note}
Each table/figure has its own build file (`tables/tbl-01-example.R`,
`figures/fig-01-example.R` - own packages and styling code, nothing
shared) and its own thin include partial (`tables/_tbl-01-example.qmd`,
`figures/_fig-01-example.qmd`) that pulls the built object in via an
include shortcode. The `export_tables`/`export_figures` targets in
`_targets.R` write that same built object to a standalone docx/TIFF in
`submission/`, not `output/` - Quarto's website render deletes anything
under `output/` it doesn't recognize as its own - so the embedded and
standalone versions are never two different definitions of the same
table/figure.
:::


::: {.content-hidden when-format="html"}
# Supplementary Information

Because notebooks, etc. are provided via the sidebar, they don't show up in HTML or Word versions. So we hide them when the format is HTML.

```{r}
#| label: si-links
#| output: asis

# Auto-generated from whatever's in supplementary/ - add a notebook there
# and it appears here with no manual list to maintain. Excludes
# notebook-template.qmd (a starting point, not content). Docx/typst only:
# hidden for html, where the sidebar nav covers it instead.
si_files <- list.files(here("manuscript/supplementary"), pattern = "\\.qmd$")
si_files <- sort(setdiff(si_files, "notebook-template.qmd"))

if (!is_interactive && !is_html) {
  for (f in si_files) {
    title <- rmarkdown::yaml_front_matter(here(
      "manuscript/supplementary",
      f
    ))$title
    base <- tools::file_path_sans_ext(f)
    cat(sprintf("- [%s](supplementary/%s.%s)\n", title, base, ext))
  }
}
```

:::


# References

::: {#refs}
:::
