---
# Render this project via `targets::tar_make()` (the `render_manuscript`
# target in _targets.R), not `quarto render manuscript.qmd` directly — this
# file reads pipeline outputs with tar_read()/tar_load(), so a direct
# render can use stale or missing data if the pipeline hasn't been run.
title: "Manuscript Title"
author:
  - name: "Your Name"
    affiliation: "Your Institution"
date: today
abstract: |
  Write a short abstract of the manuscript here.
---

# Introduction {#sec-intro}

Write your introduction here, citing sources as needed [@example2024].

```{r}
#| label: setup
#| include: false
#| warning: false

library(targets)
library(here)
here::i_am("manuscript/manuscript.qmd")
source(here("R/functions.R"))
```

```{r}
#| label: get-render-environment
#| include: false

# detect_render_format() (R/functions.R) always returns real, non-NULL
# values for every field - see that function's roxygen comment for why a
# hand-rolled version of this lookup previously broke with
# "object 'is_html' not found" whenever QUARTO_EXECUTE_INFO was unset
# (tarchetypes' dependency-scanning pass, or a standalone/interactive knit).
fmt <- detect_render_format()
is_html <- fmt$is_html
is_typst <- fmt$is_typst
is_docx <- fmt$is_docx
is_interactive <- fmt$is_interactive
ext <- fmt$ext
```

# Methods

We can add full-line equations with `\$\$` and reference them [@eq-poisson]. This uses... LatEX? Mathjax? Note that inline code fences don't get any special styling in Word. 

$$
p(x) = \frac{e^{-\lambda} \lambda^{x}}{x !}
$$ {#eq-poisson}

# Results

You can refer back to sections [@sec-intro] We had to put `{#sec-intro}` after "Introduction" in the header above.

This is a figure @fig-01-example.

{{< include figures/_fig-01-example.qmd >}}

This is a table @tbl-01-example. For some reason, they use different caption variables.

{{< include tables/_tbl-01-example.qmd >}}

::: {.callout-note}
Each table/figure has its own build file (`tables/tbl-01-example.R`,
`figures/fig-01-example.R` - own packages, own styling code, nothing
shared) and its own thin include partial (`tables/_tbl-01-example.qmd`,
`figures/_fig-01-example.qmd`) that pulls the built object in here via an
include shortcode. The `export_tables`/`export_figures` targets in
`_targets.R` write the *same* built object to a standalone docx/TIFF in
`submission/` (not `output/` — Quarto's website render deletes anything
under `output/` it doesn't recognize as its own, confirmed by direct
testing), so the embedded version above and the standalone submission file
are never two different definitions of the same table/figure.
:::


::: {.content-hidden when-format="html"}
# Supplementary Information

Because notebooks, etc. are provided via the sidebar, they don't show up in HTML or Word versions. So we hide them when the format is HTML.

```{r}
#| label: si-links
#| output: asis

# Auto-generated from whatever's actually in supplementary/ - add a new
# notebook there and it shows up here with no manual list to maintain.
# Excludes notebook-template.qmd (a starting point for new notebooks, not
# content). Docx/typst output only: this whole block is hidden for html
# (the sidebar nav covers it there), and `ext` below only knows how to
# build links for the two non-html formats.
si_files <- list.files(here("manuscript/supplementary"), pattern = "\\.qmd$")
si_files <- sort(setdiff(si_files, "notebook-template.qmd"))

if (!is_interactive && !is_html) {
  for (f in si_files) {
    title <- rmarkdown::yaml_front_matter(here(
      "manuscript/supplementary",
      f
    ))$title
    base <- tools::file_path_sans_ext(f)
    cat(sprintf("- [%s](supplementary/%s.%s)\n", title, base, ext))
  }
}
```

:::


# References

::: {#refs}
:::
