---
# Render via tar_make(), not quarto render directly - see CLAUDE.md.
title: "Colour: palettes, colorspace, and semantics"
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(RColorBrewer)
library(viridis)
library(scico)
library(colorspace)
library(MetBrewer)
library(paletteer)
library(here)
here::i_am("manuscript/supplementary/colour-mre.qmd")

set.seed(42)
```

{{< include _back-to-manuscript.qmd >}}

::: {.callout-tip}
# The Short Version
- ggplot2's own defaults are the weakest option tested — low-contrast
  continuous gradients, and a diverging scale that isn't neutral at its
  midpoint.
- `viridis` is a safe default, but check direction/range: pushed dark
  (`mako`, `direction = -1`) it can threaten gridline/label legibility.
- `scico` is the closest thing to a viridis for earth/ocean/water science
  specifically (Crameri's scientific colour maps) — the more field-relevant
  default for this project's domain.
- `paletteer` consolidates RColorBrewer/viridis/scico/MetBrewer behind one
  interface, but isn't a strict drop-in: `MetBrewer::Hokusai1` through
  `paletteer` silently dropped an 8th category with no warning. Check a
  palette's actual length before trusting it with more categories.
- Eyeballed "thematic" colours (water=blue, earth=brown) cluster in similar
  lightness/chroma and separate worse under CVD simulation than an
  algorithmic `qualitative_hcl` palette — anchor hue thematically, then
  force the spacing to match, and re-check under CVD simulation.
:::

# Why this notebook exists

Colour is usually the last thing decided and the first thing that goes
wrong. ggplot2's defaults are functional, not good; the packages that fix
that (RColorBrewer, viridis, scico, MetBrewer, colorspace) each solve a
different part of the problem, and none of them is a single drop-in
answer for every scale type. This notebook works through R's three
`ggplot2` colour-scale mechanisms — continuous (`scale_*_c`), discrete
(`scale_*_d`), and binned (`scale_*_b`) — comparing the same handful of
packages on each, then covers two specific techniques worth having on
hand: deriving a palette family from a base hue with `colorspace`, and
checking whether a thematically-motivated palette actually survives
colour-vision deficiency.

Findings below are from actually rendering the figures and looking at
them (including a CVD simulation, not just an assumption about what
"probably" survives it) — not received wisdom about which package is
best.

# The data

A synthetic (not `diamonds`) water-monitoring dataset, structured to make
categorical, continuous, and diverging demos all meaningful at once:
eight analytes (PFAS compounds, metals, nutrients, microplastics) sampled
across four sample matrices and four region types, with a made-up but
plausible regulatory guideline per analyte so `pct_exceedance` is a real
diverging quantity (positive = over guideline) rather than an arbitrary
centred variable.

```{r}
#| label: build-data
matrices <- c("Freshwater", "Seawater", "Groundwater", "Air/Deposition")
regions <- c("Coastal", "Upland", "Urban", "Agricultural")
analytes <- c(
  "PFOA",
  "PFOS",
  "Nitrate",
  "Lead",
  "Microplastics",
  "Chloride",
  "Phosphate",
  "Cadmium"
)
# Made-up but plausible guideline values (ug/L)
guidelines <- c(
  PFOA = 0.004,
  PFOS = 0.004,
  Nitrate = 50,
  Lead = 5,
  Microplastics = 1,
  Chloride = 250,
  Phosphate = 0.1,
  Cadmium = 3
)
# Matrix/region multipliers give the data real (if fabricated) structure
# instead of pure noise - urban/agricultural runoff runs higher, air
# deposition lowest, etc.
matrix_mult <- c(
  "Freshwater" = 1,
  "Seawater" = 0.4,
  "Groundwater" = 0.6,
  "Air/Deposition" = 0.15
)
region_mult <- c(Coastal = 0.8, Upland = 0.5, Urban = 1.8, Agricultural = 1.3)

water_quality <- tibble(
  site_id = sample(sprintf("S%03d", 1:60), 1200, replace = TRUE),
  region = sample(
    regions,
    1200,
    replace = TRUE,
    prob = c(0.3, 0.2, 0.25, 0.25)
  ),
  matrix = sample(
    matrices,
    1200,
    replace = TRUE,
    prob = c(0.4, 0.25, 0.25, 0.1)
  ),
  analyte = sample(analytes, 1200, replace = TRUE),
  date = as.Date("2023-01-01") + sample(0:730, 1200, replace = TRUE)
) |>
  mutate(
    guideline = guidelines[analyte],
    meanlog = log(guideline * matrix_mult[matrix] * region_mult[region]),
    concentration = rlnorm(1200, meanlog = meanlog, sdlog = 0.7),
    pct_exceedance = (concentration - guideline) / guideline * 100,
    analyte = factor(analyte, levels = names(sort(guidelines))),
    matrix = factor(matrix, levels = matrices)
  )
```

# Discrete (`_d`) scales: categorical identity

Concentration by analyte (log scale — these span four orders of
magnitude, from PFAS in the ng/L range to chloride in mg/L). Colour here
carries pure identity, not magnitude — the boxes could be any order and
lose no information, so the job of the palette is purely "make eight
things tell apart," including under colour-vision deficiency.

```{r}
#| label: fig-discrete-compare
#| fig-cap: "Concentration by analyte, five categorical palette approaches"
#| fig-width: 8
#| fig-height: 9
p_base <- ggplot(
  water_quality,
  aes(x = analyte, y = concentration, fill = analyte)
) +
  geom_boxplot(outlier.size = 0.5) +
  scale_y_log10() +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 40, hjust = 1)
  ) +
  labs(x = NULL, y = "Concentration (ug/L, log scale)")

p1 <- p_base + labs(title = "ggplot2 default hues")
p2 <- p_base +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "RColorBrewer 'Set2'")
p3 <- p_base +
  scale_fill_viridis_d(option = "viridis") +
  labs(title = "viridis (discrete)")
p4 <- p_base +
  scale_fill_manual(values = MetBrewer::met.brewer("Hokusai1", 8)) +
  labs(title = "MetBrewer 'Hokusai1'")
p5 <- p_base +
  scale_fill_manual(
    values = colorspace::qualitative_hcl(8, palette = "Dark 3")
  ) +
  labs(title = "colorspace qualitative_hcl 'Dark 3'")

(p1 | p2) / (p3 | p4) / (p5 | plot_spacer())
```

- **ggplot2 defaults**: the evenly-spaced hue wheel is *not* perceptually
  even — yellow-green and orange read as noticeably lower-contrast
  against a white background than the blues/purples in the same set, at
  identical lightness/chroma settings.
- **RColorBrewer `Set2`**: a genuinely solid qualitative set, muted and
  print-friendly, but hard-capped at 8 colours for this palette (`Set3`
  goes to 12) — check `RColorBrewer::brewer.pal.info` before assuming a
  Brewer qualitative palette covers your category count.
- **viridis (discrete)**: designed for *ordered* magnitude, not identity
  — using it for unordered categories works (it's colourblind-safe by
  construction) but the implied light-to-dark ordering is meaningless
  noise here. Fine in a pinch, not the first choice for pure identity.
- **MetBrewer `Hokusai1`**: curated from an actual artwork's palette —
  more visually distinctive/memorable than an algorithmically-spaced set,
  at the cost of no guarantee about perceptual spacing. Worth the same
  CVD check as anything else before relying on it (not done separately
  here — the CVD-simulation method is demonstrated later in this
  notebook on the `matrix` palette; apply it to whichever palette you
  actually ship).
- **colorspace `qualitative_hcl`**: algorithmically even lightness/chroma
  by construction (that's the point of specifying colour in HCL space
  rather than RGB/hue-wheel) — the "boring but correct" choice.

# Continuous (`_c`) scales: sequential magnitude

Mean % of guideline, by region and analyte — here colour *is* the data
(a heatmap), so palette quality has a much bigger effect on readability
than in the boxplots above.

```{r}
#| label: fig-continuous-compare
#| fig-cap: "Region x analyte mean concentration, four sequential palettes"
#| fig-width: 8.5
#| fig-height: 6
heat_data <- water_quality |>
  group_by(region, analyte) |>
  summarise(
    mean_pct_of_guideline = mean(concentration / guideline) * 100,
    .groups = "drop"
  )

p_heat_base <- ggplot(
  heat_data,
  aes(x = analyte, y = region, fill = mean_pct_of_guideline)
) +
  geom_tile(colour = "white", linewidth = 0.5) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  labs(x = NULL, y = NULL, fill = "% of\nguideline")

h1 <- p_heat_base +
  scale_fill_gradient(low = "white", high = "#08306B") +
  labs(title = "ggplot default gradient")
h2 <- p_heat_base +
  scale_fill_distiller(palette = "Blues", direction = 1) +
  labs(title = "RColorBrewer distiller 'Blues'")
h3 <- p_heat_base +
  scale_fill_viridis_c(option = "mako", direction = -1) +
  labs(title = "viridis 'mako' (reversed)")
h4 <- p_heat_base +
  scale_fill_scico(palette = "davos", direction = -1) +
  labs(title = "scico 'davos' (reversed)")

(h1 | h2) / (h3 | h4)
```

- **ggplot default gradient** (`scale_fill_gradient()` with hand-picked
  endpoints): flattest, muddiest low-contrast result of the four — the
  Coastal and Agricultural rows are hard to tell apart. This is the
  "arse" default the whole notebook exists to avoid.
- **RColorBrewer `distiller`**: clean, but Brewer sequential palettes
  interpolate a palette designed for ~9 discrete swatches, not
  continuous data — fine here, worth checking for banding artefacts on
  a smoother/larger dataset.
- **viridis `mako`, reversed**: highest-contrast of the four, which is
  exactly what surfaced a real caveat — pushed dark enough at the high
  end that the white gridlines and this document's dark-on-light theme
  both lose contrast against the Urban row. Reversing/extending a
  viridis palette isn't free of the same "check the actual rendered
  output" discipline as anything else here.
- **scico `davos`**: same perceptually-uniform design lineage as
  viridis (Fabio Crameri's scientific colour maps), and the closest
  thing to a field-standard for earth/ocean/atmospheric science
  specifically — worth defaulting to this, not viridis, for a
  water-science figure, purely for the "this is what reviewers in my
  field expect to see" reason as much as any perceptual one.

# Binned + diverging scales: magnitude with a meaningful zero

`pct_exceedance` (positive = over the regulatory guideline) is a genuine
diverging quantity, and regulatory data is routinely read in bins
("under guideline" / "1-2x" / ">2x") rather than as continuous shading —
`scale_*_b`/`scale_*_binned`/`scale_*_fermenter` all bin a continuous
variable into discrete-looking legend swatches while the underlying data
stays continuous.

```{r}
#| label: fig-diverging-compare
#| fig-cap: "Mean % over/under guideline, four binned diverging palettes"
#| fig-width: 8.5
#| fig-height: 6
p_div_base <- ggplot(
  heat_data |> mutate(pct_exceed = mean_pct_of_guideline - 100),
  aes(x = analyte, y = region, fill = pct_exceed)
) +
  geom_tile(colour = "white", linewidth = 0.5) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  labs(x = NULL, y = NULL, fill = "% over/under\nguideline")

d1 <- p_div_base +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  labs(title = "ggplot default diverging")
d2 <- p_div_base +
  scale_fill_fermenter(palette = "RdBu", direction = -1) +
  labs(title = "RColorBrewer fermenter 'RdBu' (binned)")
d3 <- p_div_base +
  scale_fill_steps2(
    low = "#2166AC",
    mid = "grey95",
    high = "#B2182B",
    midpoint = 0
  ) +
  labs(title = "scale_fill_steps2 (binned, custom hex)")
d4 <- p_div_base +
  colorspace::scale_fill_binned_diverging(palette = "Blue-Red 3") +
  labs(title = "colorspace diverging_hcl (binned)")

(d1 | d2) / (d3 | d4)
```

- **ggplot default** (`scale_fill_gradient2()` with literal `"blue"`/
  `"red"`): too saturated, and RGB-space interpolation through literal
  `"white"` doesn't stay perceptually neutral near the midpoint the way
  a true grey does — small values near zero read with a faint colour
  cast rather than genuinely blank.
- **The other three converge on nearly the same result.** Once the two
  real mistakes are fixed — don't use fully-saturated named RGB colours
  as endpoints, do use a genuinely neutral grey (not literal white) at
  the midpoint — RColorBrewer's `fermenter`, a hand-specified
  `scale_fill_steps2()`, and colorspace's `diverging_hcl` all land in
  the same visual place. **The specific package matters less than
  getting that structure right**, which matches this skill's own
  general colour rule: sequential is one hue light-to-dark, diverging is
  two hues plus a neutral midpoint, never a hue *at* the midpoint.

# Don't reinvent the wheel: `paletteer`

Everything above meant installing and remembering a different function
for each package (`scale_fill_brewer()`, `scale_fill_viridis_d()`,
`MetBrewer::met.brewer()`, `colorspace::qualitative_hcl()`, ...).
[`paletteer`](https://github.com/EmilHvitfeldt/paletteer) (Emil
Hvitfeldt) consolidates essentially all of them — RColorBrewer, viridis,
scico, MetBrewer, and several thousand more palettes from dozens of
other packages — behind one interface: `paletteer_d()`/`paletteer_c()`/
`paletteer_dynamic()` for raw colour vectors, and matching
`scale_*_paletteer_d()`/`_c()`/`_dynamic()` ggplot2 scales, all addressed
by a single `"package::palette"` string.

```{r}
#| label: fig-paletteer-equivalence
#| fig-cap: "Same MetBrewer palette used earlier, reached through paletteer instead - note the missing Chloride bar/legend swatch"
#| fig-width: 8
#| fig-height: 4
ggplot(water_quality, aes(wt = concentration, x = analyte, fill = analyte)) +
  geom_bar(position = "fill") +
  paletteer::scale_fill_paletteer_d("MetBrewer::Hokusai1") +
  coord_flip() +
  theme_minimal(base_size = 10) +
  labs(
    title = "MetBrewer's Hokusai1, via paletteer::scale_fill_paletteer_d()",
    x = NULL,
    y = NULL
  )
```

::: {.callout-warning}
## Found while testing: paletteer silently dropped the 8th category, no warning printed

That figure has a real bug in it, left in deliberately rather than
quietly fixed - **Chloride's bar and legend swatch are both blank**.
`water_quality$analyte` has 8 levels; `Hokusai1` has 7 native colours
(`length(paletteer::paletteer_d("MetBrewer::Hokusai1"))` confirms it);
`scale_fill_paletteer_d()`'s default `dynamic = FALSE` uses exactly the
native palette and leaves anything past its length unmapped (`NA` fill,
rendered as blank) — with no console warning at all. Compare this to the
very first figure in this notebook, which called
`MetBrewer::met.brewer("Hokusai1", 8)` **directly** and got all 8
colours (it interpolates by default when asked for more than the
palette's native count) - **the same nominal palette behaves differently
depending on which interface reaches it.**

Both obvious fixes were tried and both failed, informatively:

- `paletteer::paletteer_d("MetBrewer::Hokusai1", n = 8)` — hard error:
  `` `n` must be a whole number between 1 and 7, not the number 8 ``. At
  least this one fails loudly rather than silently.
- `scale_fill_paletteer_d("MetBrewer::Hokusai1", dynamic = TRUE)` —
  a *different* hard error: `Palette "MetBrewer::Hokusai1" not found`.
  `dynamic = TRUE` only works for palettes paletteer has separately
  registered as interpolation-capable, not every discrete palette in its
  collection — and the error message doesn't say that, it looks
  identical to a genuine typo in the palette name.

**Practical takeaway**: don't assume a palette reached through paletteer
extends to fit your data just because the same palette's own package
does it directly. Check `length(paletteer::paletteer_d("pkg::palette"))`
against your actual category count before trusting
`scale_*_paletteer_d()` with more levels than that, and don't reach for
`dynamic = TRUE` as a general-purpose fix — it's palette-specific, and
its own error message won't tell you that.
:::

**What this does and doesn't replace**: `paletteer` solves the
*access* problem (five packages, five function signatures, five sets of
palette names to remember) — it doesn't replace any judgement from the
rest of this notebook. A palette pulled through `paletteer` is exactly
as (un)safe for CVD, exactly as prone to a "looks fine at a glance"
midpoint problem in a diverging scale, as the same palette pulled
directly from its source package, because it *is* the same palette. Run
the same checks (this notebook's CVD-simulation section, in particular)
regardless of which interface fetched the colours.

**For browsing before committing to one**: Yan Holtz's
[R Color Palette Finder](https://r-graph-gallery.com/color-palette-finder.html),
built around `paletteer`'s ~2,700-palette collection, previews any
palette against seven real chart types, filters by categorical/
continuous and by length, and flags colourblind-friendly options — worth
using before hand-picking swatches the way the "thematic" palette below
was picked (see what that got tested against).

# Deriving coherent palette families with `colorspace::lighten()`

Sometimes the right palette isn't any single package's preset — it's a
small base palette (one hue per group that matters) *manipulated* to
add a second dimension. This is the pattern from the script that
prompted this notebook: pick one base colour per top-level group (there,
LLM provider; here, sample matrix), then use `colorspace::lighten()` to
fan each one out across a second variable (there, token limit; here,
region) while staying visually part of the same family.

```{r}
#| label: fig-lighten-derived
#| fig-cap: "One base hue per matrix (viridis), lightened by region (colorspace)"
#| fig-width: 7
#| fig-height: 5
base_hue <- setNames(
  viridis::viridis(4, option = "viridis", begin = 0.15, end = 0.75),
  matrices
)
lighten_amt <- c(Coastal = 0, Upland = 0.25, Urban = 0.5, Agricultural = 0.7)

derived <- expand.grid(matrix = matrices, region = names(lighten_amt)) |>
  mutate(
    fill_col = colorspace::lighten(base_hue[matrix], lighten_amt[region]),
    id = paste(matrix, region)
  )

ggplot(derived, aes(x = region, y = matrix, fill = id)) +
  geom_tile(colour = "white", linewidth = 1) +
  scale_fill_manual(values = setNames(derived$fill_col, derived$id)) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none") +
  labs(
    title = "One base hue per matrix, lightened by region",
    x = NULL,
    y = NULL
  )
```

- Any base palette works here (viridis, scico, a hand-picked set) —
  `lighten()` operates on hex colours directly, not on a palette
  object, so it composes with anything.
- `colorspace::darken()`, `desaturate()`, and `saturate()` follow the
  same pattern for the other direction/axis. `lighten()`/`darken()`
  operate in HCL space by default (`method = "relative"` on chroma and
  luminance together), which is why the result stays visually coherent
  per hue instead of just washing out toward grey the way naively
  blending toward white in RGB space would.
- The failure mode to watch for: too many steps on one base hue and the
  lightest/darkest ends stop being distinguishable from each other or
  from the background — four steps (as here) was comfortably fine, this
  wasn't stress-tested further.

# Thematic vs statistically-even colour: does "water is blue" hold up?

This is the real version of the question that prompted this section:
`matrix` (Freshwater/Seawater/Groundwater/Air-Deposition) has an obvious
thematic mapping — water is blue, ground is brown, air is pale — but
"obvious" and "perceptually well-separated" are different properties,
and only one of them is checkable by eye.

```{r}
#| label: fig-thematic-vs-statistical
#| fig-cap: "A hand-picked thematic palette vs an algorithmically-spaced one"
#| fig-width: 7
#| fig-height: 3
thematic <- c(
  "Freshwater" = "#4C8FBB",
  "Seawater" = "#0B4F6C",
  "Groundwater" = "#7C6A5A",
  "Air/Deposition" = "#C9D6DF"
)
statistical <- setNames(
  colorspace::qualitative_hcl(4, palette = "Dark 3"),
  matrices
)

swatch_df <- tibble(
  matrix = rep(matrices, 2),
  approach = rep(c("Thematic", "Statistical (qualitative_hcl)"), each = 4),
  col = c(thematic[matrices], statistical[matrices])
)

ggplot(swatch_df, aes(x = matrix, y = approach, fill = col)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(
    aes(label = matrix),
    size = 3,
    colour = "white",
    fontface = "bold"
  ) +
  scale_fill_identity() +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_blank()) +
  labs(
    x = NULL,
    y = NULL,
    title = "Thematic vs statistically-even matrix colours"
  )
```

Both look like reasonable, distinguishable palettes at normal vision.
The actual test is what happens under colour-vision deficiency —
`colorspace::simulate_cvd()` (via the `deutan()`/`protan()`/`tritan()`
wrappers) transforms a set of hex colours to approximate how each CVD
type perceives them, without needing a human tester:

```{r}
#| label: fig-cvd-check
#| fig-cap: "CVD simulation: thematic vs statistically-even matrix colours"
#| fig-width: 9
#| fig-height: 4
cvd_check <- function(cols) {
  list(
    normal = cols,
    deutan = colorspace::deutan(cols),
    protan = colorspace::protan(cols),
    tritan = colorspace::tritan(cols)
  )
}
them_cvd <- cvd_check(thematic)
stat_cvd <- cvd_check(statistical)

cvd_df <- bind_rows(
  lapply(names(them_cvd), function(v) {
    tibble(
      approach = "Thematic",
      vision = v,
      matrix = matrices,
      col = them_cvd[[v]]
    )
  })
) |>
  bind_rows(
    lapply(names(stat_cvd), function(v) {
      tibble(
        approach = "Statistical",
        vision = v,
        matrix = matrices,
        col = stat_cvd[[v]]
      )
    })
  ) |>
  mutate(
    vision = factor(vision, levels = c("normal", "deutan", "protan", "tritan"))
  )

ggplot(cvd_df, aes(x = matrix, y = vision, fill = col)) +
  geom_tile(colour = "white", linewidth = 1) +
  scale_fill_identity() +
  facet_wrap(~approach) +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(
    x = NULL,
    y = NULL,
    title = "CVD simulation: thematic vs statistical matrix colours"
  )
```

::: {.callout-warning}
## Found while testing: the "obvious" thematic palette separated worse than the algorithmic one

Under `deutan`/`protan` simulation, the hand-picked thematic palette's
four colours collapse toward similar dusty blue-grey/olive tones —
Freshwater and Seawater in particular become hard to tell apart. The
`qualitative_hcl` palette, picked with no thematic intent at all, keeps
noticeably better separation under the same simulation.

The reason isn't that thematic colour is inherently wrong — it's that
eyeballing "photographically plausible" hues (water-blue, earth-brown,
air-grey) tends to pick colours that already cluster in a similar
tonal/lightness range, which is exactly the dimension CVD collapses
hardest. A statistically-spaced palette doesn't have that bias because
nothing about its construction cares what the colours "mean."

**Practical takeaway, not "abandon theme": anchor hue thematically, then
explicitly force the lightness/chroma spacing** rather than trusting
eyeballed hex codes — e.g. pick each matrix's *hue* for its thematic
association, but pass all four through `colorspace::qualitative_hcl()`
(or manually set matching `l`/`c` values) so the spacing is actually
even, then re-run this same CVD check before trusting the result. Theme
gets you a starting hue; it doesn't get you a free pass on the actual
spacing check.
:::

# What the pros do: restrained highlight palettes

Data journalism (BBC, FT, and similar) mostly avoids this whole problem
by not needing many colours at once: one accent colour for the thing the
chart is actually about, everything else in a single neutral grey. Not
using a dedicated package for this (`bbplot` isn't on CRAN, and adding a
whole house-style package for one technique isn't worth the dependency
for what's a five-line pattern) — just a named colour vector where every
level except one maps to the same grey:

```{r}
#| label: fig-highlight
#| fig-cap: "Restrained highlight: one accent colour, everything else grey"
#| fig-width: 7
#| fig-height: 4
highlight_col <- setNames(
  rep("grey80", length(regions)),
  regions
)
highlight_col["Urban"] <- "#B2182B"

ggplot(
  water_quality |> mutate(is_urban = region == "Urban"),
  aes(x = date, y = concentration, colour = region, alpha = is_urban)
) +
  geom_jitter(size = 1) +
  scale_y_log10() +
  scale_colour_manual(values = highlight_col, guide = "none") +
  scale_alpha_manual(values = c(`TRUE` = 0.7, `FALSE` = 0.15), guide = "none") +
  theme_minimal(base_size = 10) +
  labs(
    title = "Urban sites run consistently higher",
    subtitle = "All other regions shown for context, de-emphasised",
    x = NULL,
    y = "Concentration (ug/L, log scale)"
  )
```

This isn't a competing option to the packages above so much as a
reminder that **not every figure needs a full categorical palette** — if
the actual finding is "one group differs from the rest," colour should
say exactly that, not carry all eight/four categories at equal visual
weight and make the reader do the comparison themselves.

# References
