---
# Render via tar_make(), not quarto render directly - see CLAUDE.md.
title: "Plotting Concentrations"
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(patchwork)
library(scales)
library(ggbeeswarm)
library(colorspace)
library(here)
here::i_am("manuscript/supplementary/concentrations-mre.qmd")

source(here("R/functions.R"))
# viridis instead of ggplot2's own defaults - see colour-mre.qmd. Most
# plots below set an explicit fill/colour anyway, so this mainly affects
# the density-overlay and dotplot-by-group plots.
set_default_palettes()
```

{{< include _back-to-manuscript.qmd >}}

::: {.callout-tip}
# The Short Version
- Concentration data is usually log-normal and spans orders of magnitude -
  plot on `scale_x_log10()`, not linear.
- No single geom works from n=10 to n=10k: point-based geoms (jitter/
  quasirandom/beeswarm/dotplot) show raw data but overplot at high n;
  density-based geoms (density/violin) scale but hide n and can imply
  false structure at low n. A density/violin + point overlay is the
  practical middle ground.
- `geom_beeswarm()` slows down past ~1,000 points/group; `geom_quasirandom()`
  scales to any n - prefer it when group sizes are mixed or unknown.
- `geom_dotplot()`'s `binwidth` is in data units, not counts - needs a
  per-group binwidth, a poor fit for comparing very differently sized groups.
- Colour a swarm/quasirandom plot with an evenly-spaced discrete palette
  (`colorspace::scale_colour_discrete_qualitative()`), not this project's
  `viridis` default - viridis reads unevenly once only 2-3 levels are
  pulled from it.
:::

# The data

A synthetic dataset built to mimic real concentration data: three groups
at genuinely different sample sizes and concentration scales, each drawn
from a true log-normal distribution (`rlnorm()`, not
`log1p(rnorm())`, which is not log-normal and understates skew).

```{r}
#| label: data
set.seed(8472)

combined <- bind_rows(
  tibble(
    x = rlnorm(10000, meanlog = log(100), sdlog = 0.5),
    group = "Large (n=10k)"
  ),
  tibble(
    x = rlnorm(1000, meanlog = log(10), sdlog = 0.8),
    group = "Mid (n=1k)"
  ),
  tibble(x = rlnorm(10, meanlog = log(1), sdlog = 1.0), group = "Small (n=10)")
) |>
  mutate(
    group = factor(
      group,
      levels = c("Small (n=10)", "Mid (n=1k)", "Large (n=10k)")
    ),
    # geom_dotplot() bins in data units, not log units - pre-transform.
    x_log10 = log10(x)
  )

theme_set(theme_bw(base_size = 9))

# Shared log10 x scale with tidy axis labels, reused across most plots below.
log_x <- list(
  scale_x_log10(labels = label_log()),
  labs(x = "Concentration")
)
```

# Density variants

Density estimates never overplot regardless of n, but a raw density curve
gives no sense of how many points it's built from - `n=10` and `n=10k`
can produce similarly smooth-looking curves.

```{r}
#| label: fig-density-overlay
#| fig-cap: "Density, overlaid - easy to compare shape, but n is invisible"
#| fig-width: 7
ggplot(combined, aes(x, fill = group, color = group)) +
  geom_density(alpha = 0.3, linewidth = 0.5) +
  log_x +
  labs(fill = NULL, color = NULL) +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-density-facet
#| fig-cap: "Density, faceted on a shared x-axis"
#| fig-width: 6
#| fig-height: 6
ggplot(combined, aes(x)) +
  geom_density(fill = "steelblue", alpha = 0.5, linewidth = 0.4) +
  facet_wrap(~group, ncol = 1, scales = "fixed") +
  log_x
```

Adding a rug restores some sense of n and shows individual points for the
small group, at the cost of the rug itself overplotting for the large one:

```{r}
#| label: fig-density-rug
#| fig-cap: "Density + rug - rug is informative at n=10, solid at n=10k"
#| fig-width: 7
ggplot(combined, aes(x, color = group)) +
  geom_density(linewidth = 0.6) +
  geom_rug(alpha = 0.4, length = unit(0.05, "npc")) +
  log_x +
  labs(color = NULL) +
  theme(legend.position = "bottom")
```

# Box, violin, and jitter

```{r}
#| label: fig-box
#| fig-cap: "Boxplot - compact, but hides multimodality and n"
p_box <- ggplot(combined, aes(x, y = group)) +
  geom_boxplot(outlier.size = 0.5) +
  log_x +
  labs(y = NULL)

p_box
```

```{r}
#| label: fig-violin
#| fig-cap: "Violin (scale='width') - shape at a glance, same n-blindness as density"
p_viol <- ggplot(combined, aes(x, y = group)) +
  geom_violin(scale = "width") +
  log_x +
  labs(y = NULL)

p_viol
```

A plain jittered strip shows every point, but overplots badly once n gets
into the thousands:

```{r}
#| label: fig-strip
#| fig-cap: "Jitter strip - true to the data at small n, a smear at large n"
ggplot(combined, aes(x, y = group)) +
  geom_jitter(height = 0.25, alpha = 0.15, size = 0.6) +
  log_x +
  labs(y = NULL)
```

## Combining box/violin with the raw points

Layering a box or violin with the underlying points keeps both the
summary shape and a sense of n, and reads reasonably at every group size
tested here:

```{r}
#| label: fig-box-strip
#| fig-cap: "Box + jitter - outlier points hidden on the box to avoid doubling them up"
#| fig-width: 7
p_box_strip <- ggplot(combined, aes(x, y = group)) +
  geom_jitter(height = 0.25, alpha = 0.15, size = 0.6, color = "grey50") +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  log_x +
  labs(y = NULL)

p_box_strip
```

```{r}
#| label: fig-violin-box
#| fig-cap: "Violin + box, nested - shape and quartiles together"
#| fig-width: 7
p_viol_box <- ggplot(combined, aes(x, y = group)) +
  geom_violin(scale = "width", fill = "steelblue", alpha = 0.4) +
  geom_boxplot(width = 0.12, outlier.size = 0.3, fill = "white") +
  log_x +
  labs(y = NULL)

p_viol_box
```

# Dotplots

`geom_dotplot()`'s `binwidth` is in the units of the (here, already
log10-transformed) x-axis, not a point count - the same value has to be
tuned per group size or the plot either collapses to one column (too
coarse) or thins to invisibility (too fine). Below, each group gets its
own tuned `binwidth`, on its own x-range:

```{r}
#| label: fig-dotplot-tuned
#| fig-cap: "Dotplots, tuned per group - readable individually, not directly comparable"
#| fig-width: 6
#| fig-height: 7
dot_scale <- scale_x_continuous(
  labels = \(b) label_log()(10^b),
  name = "Concentration"
)

p_dot_small <- ggplot(filter(combined, group == "Small (n=10)"), aes(x_log10)) +
  geom_dotplot(binwidth = 0.12, stackdir = "center", alpha = 0.7, stroke = 0) +
  dot_scale +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = "Small (n=10)")

p_dot_mid <- ggplot(filter(combined, group == "Mid (n=1k)"), aes(x_log10)) +
  geom_dotplot(binwidth = 0.03, stackdir = "center", alpha = 0.5, stroke = 0) +
  dot_scale +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = "Mid (n=1k)")

p_dot_big <- ggplot(filter(combined, group == "Large (n=10k)"), aes(x_log10)) +
  geom_dotplot(binwidth = 0.005, stackdir = "center", alpha = 0.3, stroke = 0) +
  dot_scale +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = "Large (n=10k)")

p_dot_small / p_dot_mid / p_dot_big
```

Forcing one `binwidth` and one shared x-range across all three groups
makes them directly comparable, but the large group loses individual
dots entirely - it reads as a solid bar rather than n=10,000 points:

```{r}
#| label: fig-dotplot-shared
#| fig-cap: "Dotplots, faceted on a common x-axis and binwidth"
#| fig-width: 6
#| fig-height: 6
x_lims <- range(combined$x_log10) + c(-0.05, 0.05)

ggplot(combined, aes(x_log10)) +
  geom_dotplot(binwidth = 0.04, stackdir = "center", alpha = 0.6, stroke = 0) +
  facet_wrap(~group, ncol = 1, scales = "fixed") +
  scale_x_continuous(
    labels = \(b) label_log()(10^b),
    name = "Concentration",
    limits = x_lims
  ) +
  scale_y_continuous(NULL, breaks = NULL) +
  coord_transform(clip = "off")
```

**Takeaway: dotplots are most useful for a single small/mid-sized group,
or a handful of groups of similar n shown side by side** - they stop
being informative once one group's n is orders of magnitude larger than
another's on the same panel.

# Beeswarm and quasirandom

`ggbeeswarm` provides two point-jitter strategies: `geom_quasirandom()`
(a van der Corput sequence - deterministic, density-proportional, not
strictly non-overlapping) and `geom_beeswarm()` (strict non-overlap,
visibly slower and impractical much above ~1,000 points per group).

Colouring points by group needs a discrete scale that stays perceptually
even at whatever group count you have - this project's own default
(`set_default_palettes()`, `viridis`) is designed to be read along its
full gradient, so pulling just 2-3 discrete levels from it grabs uneven
endpoints (a dark purple next to a pale yellow), which reads unevenly
once alpha-blended over overlapping points.
`colorspace::scale_colour_discrete_qualitative()` picks hues at matched
lightness/chroma instead - the same fix `colour-mre.qmd` recommends for
categorical colour generally:

```{r}
#| label: fig-quasirandom
#| fig-cap: "Quasirandom - scales to any n, denser regions read as denser"
p_quasi <- ggplot(combined, aes(x, y = group, colour = group)) +
  geom_quasirandom(orientation = "y", alpha = 0.3, size = 0.75) +
  scale_colour_discrete_qualitative() +
  log_x +
  labs(y = NULL, colour = NULL) +
  theme(legend.position = "bottom")

p_quasi
```

```{r}
#| label: fig-quasirandom-box
#| fig-cap: "Quasirandom + box"
#| fig-width: 7
p_quasi_box <- ggplot(combined, aes(x, y = group)) +
  geom_quasirandom(
    orientation = "y",
    alpha = 0.15,
    size = 0.5,
    color = "grey40"
  ) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  log_x +
  labs(y = NULL)

p_quasi_box
```

`geom_beeswarm()` restricted to the small + mid groups only - confirmed
by direct testing (and SAW) that including the n=10k group here makes the render
noticeably slower (and basically uninterpretable), consistent with the "impractical above ~1k points"
guidance in the package's own documentation:


```{r}
#| label: fig-beeswarm-smallmid
#| fig-cap: "Beeswarm, small + mid groups only - strict non-overlap, clearest shape at low n"
p_bee_smallmid <- ggplot(
  filter(combined, group != "Large (n=10k)"),
  aes(x, y = group, colour = group)
) +
  geom_beeswarm(orientation = "y", alpha = 0.5, size = 1, cex = 1.5) +
  scale_colour_discrete_qualitative() +
  log_x +
  labs(y = NULL, colour = NULL) +
  theme(legend.position = "bottom")

p_bee_smallmid
```

# References
