---
# Render via `targets::tar_make()` (the `render_manuscript` target in
# _targets.R), not `quarto render` directly - see manuscript.qmd.
#
# Minimal reproducible example (MRE) for diagram-drawing tools in Quarto,
# beyond the mermaid coverage already in images-mre.qmd: Graphviz/DOT (the
# knitr `dot` engine and DiagrammeR::grViz), cetz (typst-native drawing,
# reached via a raw `{=typst}` block), and grammar-of-graphics network
# diagrams in R (ggraph). Render all three formats and open them next to
# each other.
#
# To surface this notebook in the site navigation, see the comment in
# supplementary/notebook-template.qmd.
title: "Diagrams in Quarto: Graphviz, cetz, and ggraph"
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(ggraph)
library(tidygraph)
library(DiagrammeR)
library(here)
here::i_am("manuscript/supplementary/diagrams-mre.qmd")
source(here::here("R/functions.R"))
fmt <- detect_render_format()

set.seed(1)
```

{{< include _back-to-manuscript.qmd >}}

::: {.callout-tip}
# The Short Version
- **Mermaid** already covers the "zero-install flowchart" case — see
  [images-mre.qmd](images-mre.qmd), not repeated here.
- **Graphviz's `dot` engine** shells out to a system `dot` binary — and it
  actually worked here, but only found via whatever environment the real
  `tar_quarto()` render process resolves, not the shell used to sanity-check
  it beforehand. See the callout below — this is a live instance of the
  "don't trust a shell's PATH without checking the process that matters"
  finding already in CLAUDE.md's Pipeline gotchas.
- **`DiagrammeR::grViz()`** draws the same DOT language via a bundled JS
  library instead (no system binary needed) — but that makes it an
  HTML/JS widget, same category as `DT`/`visNetwork` in
  [tables-mre.qmd](tables-mre.qmd): fine in html, not something
  docx/typst can render directly.
- **cetz** rendered correctly here, canvas and all — contradicting an
  open upstream report that canvas-wrapped diagrams fail. See the
  callout below for exact versions and what that means (and doesn't
  mean) for trusting it.
- **ggraph** (tidygraph + ggplot2) produces an ordinary ggplot object, so
  it inherits this project's existing ggplot findings
  ([images-mre.qmd](images-mre.qmd), [plots-mre.qmd](plots-mre.qmd))
  rather than being a new format-support question.
:::

# Why this notebook exists

Flowcharts and network diagrams show up constantly in manuscripts, and
"just use Mermaid" (see [images-mre.qmd](images-mre.qmd)) doesn't cover
every case — hand-drawn technical diagrams (Graphviz/DOT), typst-native
vector drawing (cetz), and actual data-driven network graphs (ggraph) are
different tools for different jobs. Findings below are from direct
testing in this template against `--to html`, `--to docx`, and
`--to typst`, not assumed from each tool's own documentation.

# Graphviz / DOT

## The knitr `dot` engine

Quarto/knitr can execute a ```` ```{dot} ```` chunk directly, but unlike
Mermaid this shells out to a **system** `dot` executable rather than
bundling anything — Graphviz has to be separately installed and
resolvable on `PATH` at chunk-execution time.

```{dot}
//| label: fig-dot-engine
//| fig-cap: "A DOT graph via the knitr dot engine"
//| error: true
digraph {
  rankdir = LR
  A [label = "Raw data"]
  B [label = "Processing"]
  C [label = "Manuscript"]
  A -> B -> C
}
```

`error: true` kept above as a defensive habit (same reasoning as gating
any other format-specific failure in this template) even though it
turned out not to be needed — see the callout below.

::: {.callout-warning}
## Found while testing: a clean `Sys.which("dot")` check doesn't mean what you'd think

This chunk **rendered successfully** — real Graphviz-computed layout,
confirmed by inspecting the actual SVG markup in `output/supplementary/diagrams-mre.html`
(node ellipses, edge paths, `font-family="Times,serif"` — genuine `dot`
output, not a fallback). That's despite `Sys.which("dot")` returning
empty in a plain `Rscript -e 'Sys.which("dot")'` check run immediately
before and after the same render, in what looks like the same shell
environment.

This is a live instance of a finding CLAUDE.md's "Pipeline gotchas"
section already documents for a different symptom: a single `tar_make()`
run involves multiple separately-resolved processes (the R driving
`tar_make()`, the `quarto` CLI subprocess, and the R that `quarto` itself
spawns to execute chunks), and none of them are guaranteed to see the
same `PATH` as whatever shell you happen to be checking from. **The
practical upshot is the same as before, just for a new symptom: don't
trust a "not on PATH" finding from a diagnostic shell as proof a render
will fail — check whether the actual render fails, not whether a
side-channel check says it should.** Conversely: if a `{dot}` chunk *does*
fail for you, `Sys.which("dot")` from your own terminal isn't a reliable
way to rule out "it's actually there," or to confirm it's really missing
— check the render's own error output.
:::

## `DiagrammeR::grViz()`

Same DOT language, but rendered via `viz.js`, a JS Graphviz port bundled
inside the `DiagrammeR` package — no system `dot` install required.

```{r}
#| label: fig-grviz
#| fig-cap: "The same graph via DiagrammeR::grViz()"
#| eval: !expr fmt$is_html
grViz("
digraph {
  rankdir = LR
  A [label = 'Raw data']
  B [label = 'Processing']
  C [label = 'Manuscript']
  A -> B -> C
}
")
```

::: {.content-visible unless-format="html"}
*`DiagrammeR::grViz()` output omitted here — it's an htmlwidget with no
built-in static export, same limitation class as `DT`/`visNetwork` in
[tables-mre.qmd](tables-mre.qmd). `DiagrammeR::export_svg()` plus a
headless-browser screenshot step (`rsvg`/`webshot2`) can produce a static
image for docx/typst, but that's a separate rendering pipeline, not
something Quarto does automatically the way it does for Mermaid.*
:::

# cetz: typst-native vector drawing

[cetz](https://github.com/cetz-package/cetz) is typst's most popular
drawing package — the closest thing typst has to TikZ. Quarto has no
built-in cetz support; the only way in is a raw `` {=typst} `` block that
imports the package directly, which only typst output actually uses —
pandoc drops raw `{=typst}` content silently for html/docx.

```{=typst}
#import "@preview/cetz:0.3.4"
#cetz.canvas({
  import cetz.draw: *
  circle((0, 0), radius: 1)
  line((0, 0), (2, 2))
  content((2.3, 2), [Manuscript])
})
```

::: {.content-visible unless-format="typst"}
*Nothing renders above in html/docx — raw `{=typst}` content is typst-only
by design. See the callout below for what happened in the typst render.*
:::

::: {.callout-note}
## Found while testing: the reported canvas failure did NOT reproduce here

There's an open upstream report
([quarto-dev/quarto-cli#12439](https://github.com/quarto-dev/quarto-cli/discussions/12439))
that any diagram wrapping a `cetz.canvas` fails to render through Quarto,
attributed to how Quarto assembles the typst preamble. The block above
uses `cetz.canvas` deliberately, to test that exact failure mode rather
than a simplified case that dodges it — **it rendered correctly**:
`quarto call typst-gather`-style `@preview` fetch pulled `cetz:0.3.4` and
its `oxifmt:0.2.1` dependency automatically at render time, and the
circle/line/"Manuscript" label all show up in the compiled PDF — confirmed
by `pdftotext`-extracting `output/supplementary/diagrams-mre.pdf` and
finding the literal "Manuscript" text label from the `content(...)` call,
not just a successful exit code.

Tested with Quarto 1.10.18. Don't read this as "the report was wrong" —
open discussions like that one are often narrower than their title
(a specific extension/preamble combination, an older Quarto/typst
version, a more complex canvas) and this environment may simply not hit
whatever specific trigger they found. Treat cetz as **worth trying, not
assumed-broken** — but given it needs a live fetch from the `@preview`
package registry the first time (no vendoring here, unlike the
`typst-gather` option for custom Quarto formats — see
[Custom Typst Formats](https://quarto.org/docs/output-formats/typst-custom.html)),
also don't assume it'll work offline or in a network-restricted CI
runner without that extra step.
:::

# Grammar-of-graphics network diagrams: ggraph

`ggraph` (built on `tidygraph`) draws graphs the same way `ggplot2` draws
everything else — layers, aesthetics, a `+`-composed spec — rather than a
separate DSL like DOT or Mermaid. Because the output is an ordinary
`ggplot` object, it should inherit this project's existing ggplot
findings (device, fonts, sizing — see
[images-mre.qmd](images-mre.qmd)/[plots-mre.qmd](plots-mre.qmd)) rather
than raising new format-support questions of its own.

```{r}
#| label: fig-ggraph
#| fig-cap: "A small directed graph via ggraph"
g <- tidygraph::tbl_graph(
  nodes = data.frame(name = c("Raw data", "Processing", "Manuscript")),
  edges = data.frame(from = c(1, 2), to = c(2, 3))
)

ggraph(g, layout = "linear") +
  geom_edge_link(
    arrow = arrow(length = unit(3, "mm")),
    end_cap = circle(4, "mm")
  ) +
  geom_node_point(size = 6, colour = "steelblue") +
  geom_node_text(aes(label = name), vjust = -1.5) +
  theme_graph()
```

- ✅ Renders like any other ggplot figure — same device (`ragg_png`),
  same `fig-width`/`fig-height`/`fig-dpi` defaults, no format-specific
  handling needed.
- The right tool when a diagram's layout should be *computed from data*
  (an actual graph/network structure) rather than hand-positioned, which
  is what Mermaid/DOT/cetz all require.

# References
