---
# Render via tar_make(), not quarto render directly - see CLAUDE.md.
title: "Fonts in ggplot2: sources and renderers"
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(ggtext)
library(emoji)
library(here)
here::i_am("manuscript/supplementary/fonts-mre.qmd")
```

{{< include _back-to-manuscript.qmd >}}

::: {.callout-tip}
# The Short Version
- Two separate font systems: document-level fonts (html/docx/typst
  `mainfont` - already in CLAUDE.md) govern prose; figure-level fonts are
  resolved by whichever R graphics device drew the plot.
- **A real reproducibility gap**: Fira Sans/Code (used throughout this
  project) are installed on this dev machine but nothing in the repo
  installs them - works today by accident, not by design.
- For a local/licensed font file, `systemfonts::register_font()` +
  `ragg::agg_png()` renders it correctly; `showtext`+`font_add_google()`
  covers web fonts but needs `showtext_opts(dpi=)` to match the device
  DPI. `extrafont::font_import()` hard-errors on anything outside system
  font directories. `svglite` references a registered font by internal
  name and doesn't embed it.
- **Root-caused the `plots-mre.qmd` warning**: `font family not found in
  Windows font database` is base `png()`'s legacy GDI lookup failing -
  fixed project-wide via `knitr: opts_chunk: dev: "ragg_png"`.
- Colour emoji glyphs render correctly through `ragg::agg_png()` - but a
  whole-string font override (e.g. `element_text(family = "Noto Emoji")`)
  on mixed emoji+text wrecks the Latin spacing; isolate the emoji to its
  own quoted `ggtext` span instead.
:::

# Why this notebook exists

[plots-mre.qmd](plots-mre.qmd) hit a real warning
(`font family not found in Windows font database`) while combining a
`flextable` grob with a plot. That warning turned out to be a small
symptom of a much bigger question: fonts in ggplot2 figures are resolved
by a completely different mechanism than the document-level fonts
CLAUDE.md already covers, several competing R packages exist to manage
them, each with its own eccentricities, and more than one graphics device
might end up rendering any given figure. Findings below are from direct
testing in this template - every claim here was actually reproduced, not
assumed from a package's own documentation.

# Two font systems, not one

**Document-level fonts** (html `mainfont`/`monofont`, the docx
reference-doc's theme font, typst `mainfont` - see CLAUDE.md) govern
prose: headings, body text, captions. Pandoc/typst/Word resolve these
themselves, at the document level, once per render.

**Figure-level fonts** are different: a `ggplot2` figure is rendered to a
PNG/SVG by an R *graphics device* before pandoc ever sees it, and that
device has to resolve any font name used inside the plot (via `theme()`,
`element_text(family = ...)`, etc.) on its own. A plot's embedded text has
already been turned into pixels or vector paths by the time pandoc
touches the document - nothing about `mainfont`/`monofont` reaches inside
a figure. This is why "the document renders in the right font, but the
plot's own title doesn't" is a completely different problem from anything
already documented, and needs its own fix.

# The reproducibility gap hiding in plain sight

```{r}
#| label: check-installed-fonts
library(systemfonts)
fonts <- system_fonts()
cat(
  "Fira Sans installed on this machine:",
  any(grepl("Fira Sans", fonts$family)),
  "\n"
)
cat(
  "Fira Code installed on this machine:",
  any(grepl("Fira Code", fonts$family)),
  "\n"
)
```

Both are installed here - confirmed above, not assumed - which is exactly
why using `family = "Fira Sans"` inside a `ggplot2` theme has never
visibly broken during this project's local development. But nothing in
`runme.R`, `_targets.R`, or `.github/workflows/render.yml` installs these
fonts anywhere. A fresh Ubuntu CI runner or a new contributor's machine
has no reason to have them either - the only reason this hasn't already
surfaced as a visible bug is that no figure in this project's actual
manuscript currently sets an explicit font family. The moment one does,
whether it renders correctly becomes a matter of luck about whose machine
it's rendered on, not something this repository guarantees.

# Registering a local font file: systemfonts + ragg

The realistic scenario for a future brand font isn't "add a Google Fonts
entry" - it's "someone hands you a licensed `.otf`/`.ttf` file." A fixture
font (`manuscript/fonts/SpaceMono-Regular.ttf`, generated once by
[`generate-fonts.R`](../generate-fonts.R) and committed to git, the same
pattern `generate-images.R` uses) stands in for that case here - chosen
because it's visually distinctive (a geometric monospace) and confirmed
**not** preinstalled on this machine, so a correct render actually proves
registration worked rather than accidentally matching a font that was
there anyway.

```{r}
#| label: fig-local-font-file
#| fig-cap: "A font registered from an arbitrary local file, not a system font"
register_font(
  name = "Space Mono Local",
  plain = here("manuscript/fonts/SpaceMono-Regular.ttf")
)

ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(title = "Space Mono, registered from a local file") +
  theme_minimal(base_family = "Space Mono Local")
```

No system-wide font install, no admin rights, no OS-specific step - just
a file path. This is confirmed to work with `ragg::agg_png()` (this
project's default figure device, see below) as of `systemfonts` 1.3.2 and
`ragg` 1.5.2.

# Remote web fonts: showtext + sysfonts::font_add_google()

`sysfonts::font_add_google()` fetches a font directly from Google's live
API - genuinely useful for "I just want a Google Font, no file to manage
myself," but it's a real network dependency at render time, unlike the
committed fixture file above. `showtext_auto()` then makes `showtext`
intercept all subsequent text drawing for the rest of this session -
turned back off further down once the point's made.

```{r}
#| label: fig-showtext-google-font
#| fig-cap: "A live Google Font (Space Mono again, fetched remotely this time) via showtext"
library(sysfonts)
library(showtext)

font_add_google("Space Mono", "spacemono_remote")
showtext_auto()

p_showtext <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(title = "Space Mono, fetched live via font_add_google()") +
  theme_minimal(base_family = "spacemono_remote", base_size = 14)
p_showtext
```

## The bug: showtext's DPI has to match the device's, or text is the wrong size

`showtext_opts()` defaults to 96 DPI regardless of what DPI the actual
graphics device renders at. This project's figures render at 300 DPI
(`fig-dpi: 300` in `manuscript/_quarto.yml`) - if `showtext_opts(dpi = )`
is never told that, `showtext`'s text comes out sized for a 96 DPI device
squeezed into a 300 DPI canvas, visibly too small relative to every other
plot element:

```{r}
#| label: fig-showtext-dpi-mismatch
#| fig-cap: "showtext at the default 96 DPI, rendered on a 300 DPI device - text is too small"
#| fig-dpi: 300
p_showtext
```

```{r}
#| label: fig-showtext-dpi-fixed
#| fig-cap: "Same plot, showtext_opts(dpi = 300) matching the actual device - correct size"
#| fig-dpi: 300
showtext_opts(dpi = 300)
p_showtext
```

Confirmed side by side: the fix isn't in the plot code at all, it's
`showtext_opts(dpi = )` matching whatever DPI the device actually renders
at - easy to get wrong once, and easy to forget applies per-document, not
globally, if a project renders at more than one DPI.

```{r}
#| label: showtext-cleanup
#| include: false
# Turn showtext back off - it intercepts every text-drawing call for the
# rest of this R session otherwise, which would affect every chunk below.
showtext_auto(FALSE)
```

# svglite handles registered fonts differently

`svglite` is a vector device (already used elsewhere in this project for
`sample-diagram.svg`, see [images-mre.qmd](images-mre.qmd)) - worth
checking separately from `ragg`'s raster output, since vector and raster
devices don't have to resolve fonts the same way.

```{r}
#| label: check-svglite-font-reference
svglite_path <- tempfile(fileext = ".svg")
p_local <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(title = "Space Mono Local, drawn by svglite") +
  theme_minimal(base_family = "Space Mono Local")
svglite::svglite(svglite_path, width = 8, height = 5)
print(p_local)
dev.off()

svg_lines <- readLines(svglite_path, warn = FALSE)
font_refs <- unique(regmatches(
  svg_lines,
  regexpr('font-family: "[^"]*"', svg_lines)
))
font_refs <- font_refs[nzchar(font_refs)]
cat("What the SVG actually references:", font_refs[1], "\n")
```

Confirms a real difference from `ragg`, not a rendering artifact:
`svglite` writes `font-family: "Space Mono"` - the font's own internal
family name, from the file's metadata - not `"Space Mono Local"`, the
alias `register_font()` was given it under. More consequential than the
naming mismatch itself: the SVG only *references* a font by name, it
doesn't embed the glyphs the way a raster PNG bakes them into pixels.
Opening this SVG on a machine without "Space Mono" (the real name)
available falls back to whatever the browser/viewer picks instead - a
`ragg`-rendered PNG has no equivalent failure mode, since the actual
pixels are already drawn.

# extrafont: why it doesn't fit this project

`extrafont::font_import()` is built around scanning *system* font
directories (`C:\Windows\Fonts` and similar) - not an arbitrary path.
Pointed at the same local file used successfully above:

```{r}
#| label: extrafont-failure
#| error: true
library(extrafont)
font_import(paths = here("manuscript/fonts"), prompt = FALSE)
```

That's not "imports nothing gracefully" - it's a hard error, confirmed
directly, because `extrafont`'s internal code doesn't expect a pattern
match to come back empty. To use `extrafont` at all here, the font would
first need a full, separate, OS-level font installation step outside R
entirely - real friction `systemfonts::register_font()` simply doesn't
have, and irrelevant to how this project's figures actually render
(`ragg`/`svglite`, not the base `pdf()`/`postscript()` devices
`extrafont` was really built around).

# Root-causing (and fixing) the gen_grob() warning

The warning that started this notebook, reproduced directly by forcing
the *base* `png()` device instead of this project's default:

```{r}
#| label: fig-default-device-warning
#| fig-cap: "flextable::gen_grob() output via the base png() device - produces the warning"
#| dev: "png"
#| warning: true
library(flextable)
library(patchwork)

ft <- flextable(head(mtcars[, c("mpg", "wt", "cyl")], 5)) |>
  theme_vanilla() |>
  flextable::bold(part = "header") |>
  autofit()
tbl_grob <- flextable::gen_grob(ft, fit = "fixed", scaling = "min")
p_a <- ggplot(mtcars, aes(wt, mpg)) + geom_point() + theme_minimal()

p_a + wrap_elements(full = tbl_grob)
```

```{r}
#| label: fig-ragg-device-no-warning
#| fig-cap: "Identical figure, this project's actual default device (ragg_png) - no warning"
#| warning: true
p_a + wrap_elements(full = tbl_grob)
```

Confirmed: the warning is specific to the base `png()` device, which on
Windows resolves fonts through the legacy GDI font database (the exact
mechanism named in the warning text itself) - `ragg::agg_png()`, which is
`systemfonts`-aware, resolves the identical font with nothing printed.
**Fixed project-wide, not just documented**: `manuscript/_quarto.yml` now
sets `knitr: opts_chunk: dev: "ragg_png"`, so every figure in this
project - not just this notebook's own chunks - renders through the
device confirmed to handle fonts correctly, including the local-file
registration case above that this project would actually need for a
future brand font.

# Other web font providers

Not tested directly, but worth knowing about if brand-matching comes up
later:

- **Bunny Fonts** - a drop-in mirror of the same Google Fonts catalog,
  EU-hosted, no Google tracking. Same integration pattern as Google Fonts
  (self-host the files, or point at their CDN) - worth it specifically
  for GDPR-conscious institutional use, not because the fonts differ.
- **Adobe Fonts (Typekit)** - a real caveat, not just an alternative:
  licensing normally ties it to Creative Cloud desktop sync, not
  downloadable static files, so it typically doesn't fit a reproducible
  pipeline like this one at all unless the specific font is separately
  licensed as a static file.

# Emoji: a first pass

An obvious next step from "fonts in ggplot2" - can a plot use icons the
way a good piece of game UI does (*Civilization VI*'s yield/resource
icons are a genuinely nice example: legible at a glance, consistent
weight, sit inline with numbers)? Two real problems stand between "type
an emoji" and that:

1. Rendering an emoji **as the data mark itself** (`geom_point`-as-emoji)
   is supported by more than one package - and almost never what you
   actually want. Demonstrated, not just asserted, below.
2. Icons used *properly* - inline with text, with real control over
   size/margin/baseline - need either a font that behaves (most emoji
   fonts don't, for ordinary Latin text sharing the same run) or an
   actual image, which sidesteps font metrics entirely at the cost of
   needing a real asset per icon instead of typing a character.

This is a first pass at the package/font landscape, not a solved
design system - scoped to "does the obvious thing work, and where does
it break."

## Does a colour emoji even render? (Yes - the harder question is what breaks around it)

```{r}
#| label: check-emoji-fonts
droplet <- emoji::emoji("droplet")
cat("Emoji-capable fonts on this machine:\n")
print(systemfonts::system_fonts()$family[
  grepl("emoji|symbol", systemfonts::system_fonts()$family, ignore.case = TRUE)
])
```

```{r}
#| label: fig-emoji-colour-whole-title
#| fig-cap: "Colour emoji font, whole title set to that family"
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(title = paste0(droplet, " Water yield (colour font, whole title)")) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(family = "Segoe UI Emoji"))
```

That renders correctly - a real colour droplet, ordinary-looking title
text around it. **Not the safe assumption it looks like**: most R
graphics devices historically couldn't draw colour (COLR/CBDT-format)
emoji fonts at all, only monochrome outline glyphs. `ragg::agg_png()`
(this project's device - see above) handles it fine, confirmed here, not
assumed from `ragg`'s own documentation.

## The actual footgun: whole-string family overrides don't degrade gracefully

Swap the same approach to a *monochrome* emoji font (`"Noto Emoji"`,
distinct from `"Noto Color Emoji"` - both are installed here, confirmed
above):

```{r}
#| label: fig-emoji-mono-whole-title
#| fig-cap: "Monochrome emoji font, whole title set to that family - breaks the Latin text too"
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(title = paste0(droplet, " Water yield (monochrome font, whole title)")) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(family = "Noto Emoji"))
```

The ordinary words in the title are visibly, badly mis-spaced - `"Noto
Emoji"` has no real Latin letterforms to fall back on, so the *whole*
title's layout breaks, not just the glyph. `"Segoe UI Emoji"` above
didn't have this problem because it's an OS-integrated font with its own
sane fallback chain; a standalone downloaded emoji-only font isn't
guaranteed the same courtesy. Setting `family` for an entire label is
fine when it's genuinely one font's job (as everywhere else in this
notebook) and a real risk the moment that label mixes an emoji with
ordinary prose.

The fix: isolate the emoji to its own `ggtext` span instead of setting
`family` for the whole label.

```{r}
#| label: fig-emoji-mono-isolated
#| fig-cap: "Same monochrome font, isolated to a ggtext span - Latin text unaffected"
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(
    title = paste0(
      "<span style=\"font-family:'Noto Emoji'\">",
      droplet,
      "</span> ",
      "Water yield (isolated span, monochrome)"
    )
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = ggtext::element_markdown())
```

::: {.callout-warning}
## Found while testing: an unquoted CSS font-family name fails silently, not loudly

The first version of the chunk above used
`style="font-family:Noto Emoji"` (no quotes around the two-word family
name) - it rendered *without error*, but the glyph came back in colour,
not the requested monochrome font. `gridtext`/`ggtext`'s markdown-to-CSS
parsing didn't match the unquoted multi-word name, silently fell back to
something else, and R's normal font-substitution machinery picked a
colour-capable font as the substitute rather than failing visibly.
**Quoting the family name
(`font-family:'Noto Emoji'`, single quotes inside the double-quoted HTML
attribute) fixed it** - confirmed side by side, same code otherwise. A
font-family mismatch here doesn't look like a bug; it looks like a
plausible result with the wrong font, which is worse.
:::

## Package options

- **`emoji`** (used above via `emoji::emoji("droplet")`) - a name→unicode
  lookup table, nothing more. It doesn't render anything itself; it just
  saves hunting for the raw code point. Names follow the standard
  Unicode CLDR short-name convention.
- **`ggtext`** (already used elsewhere in this project - see
  [plots-mre.qmd](plots-mre.qmd)) - lets a label mix fonts within one
  string via markdown/HTML spans, which is the actual mechanism that
  makes isolating an emoji's font practical, per the callout above.
- **`emojifont`** - an older package with its own `geom_emoji()` and its
  own internal emoji-name registry, **not the same names as the `emoji`
  package**:

```{r}
#| label: fig-emojifont-failure
#| fig-cap: "emojifont::geom_emoji() with a valid, current Unicode short name"
#| error: true
ggplot(mtcars[1:5, ], aes(wt, mpg)) +
  emojifont::geom_emoji("sheaf_of_rice", size = 0.05) +
  theme_minimal(base_size = 14)
```

  `"sheaf_of_rice"` is a real, current Unicode CLDR name (it's what
  `emoji::emoji()` uses successfully elsewhere in this section) -
  `emojifont` doesn't recognise it, confirmed by the error above, not
  assumed from its age. The two packages' emoji-name vocabularies aren't
  interchangeable; picking one commits to its naming scheme.
- **`ggimage`** (installed, **not exercised in this pass**) - treats an
  icon as an actual image grob (`geom_image()`) rather than a font
  glyph, which is the real answer to this notebook's problem 2 (font
  typesetting doesn't give real control over icon margins/baseline/size
  the way an image does) - at the cost of needing an actual image asset
  per icon rather than typing a character. Worth a dedicated pass of its
  own rather than folding in here.

## Why "emoji as the data mark" is rarely what you want

It does work:

```{r}
#| label: fig-emoji-as-marks
#| fig-cap: "Emoji glyphs used directly as geom_text() marks"
wheat <- emoji::emoji("sheaf_of_rice")
ggplot(mtcars[1:5, ], aes(wt, mpg)) +
  geom_text(label = wheat, family = "Segoe UI Emoji", size = 8) +
  theme_minimal(base_size = 14) +
  labs(title = "geom_text() with an emoji glyph directly")
```

Working is exactly the problem: at any size legible enough to read as an
icon, five colourful, unevenly-weighted glyphs already dominate the
plot's visual hierarchy more than the data does, they don't share a
common visual "centre" the way a circle/triangle does (so precise
position judgement suffers), and that gets worse, not better, as point
count grows. Fine for a one-off illustration; a poor default for actual
data marks, which is exactly what was suspected going in - confirmed
here rather than just asserted.

# References
