---
# Render via `targets::tar_make()` (the `render_manuscript` target in
# _targets.R), not `quarto render` directly - see manuscript.qmd.
#
# Minimal reproducible example (MRE) for fonts in ggplot2 figures
# specifically - not the document-level fonts (html mainfont/monofont,
# the docx reference-doc, typst mainfont) already covered in CLAUDE.md,
# which is a completely separate mechanism from what this notebook is
# about.
title: "Fonts in ggplot2: web fonts, local files, and three renderers"
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(here)
here::i_am("manuscript/supplementary/fonts-mre.qmd")
```

{{< include _back-to-manuscript.qmd >}}

::: {.callout-tip}
# The Short Version
- **Two entirely separate font systems, easy to conflate**: document-level
  fonts (html `mainfont`, the docx reference-doc, typst `mainfont` -
  already in CLAUDE.md) govern prose. Figure-level fonts - what this
  notebook covers - are resolved by whatever R graphics device drew the
  plot, a completely different mechanism.
- **A reproducibility gap confirmed hiding in plain sight**: Fira
  Sans/Fira Code (used throughout this project) are installed on this
  development machine, confirmed via `systemfonts::system_fonts()` - but
  nothing in this repo installs them anywhere. Nothing has broken *yet*
  only because of what happens to already be on this one machine.
- **`systemfonts::register_font()` + `ragg::agg_png()` is the recommended
  combination** - confirmed to correctly render a font registered from an
  arbitrary local *file*, not just a system-installed font. This is the
  realistic case for a future brand font: a licensed file someone hands
  you, not a Google Fonts catalog entry.
- **`showtext` + `sysfonts::font_add_google()` works for remote web
  fonts, but has a confirmed real footgun**: `showtext_opts(dpi = )` must
  match the actual device DPI, or text renders at the wrong size relative
  to the rest of the plot - confirmed by rendering the identical plot both
  ways.
- **`svglite` handles registered fonts differently, confirmed by
  inspecting the raw SVG**: it references the font by its own internal
  name, not whatever alias you registered it under, and does not embed
  it - unlike a raster PNG, the SVG isn't self-contained.
- **`extrafont` is confirmed a poor fit for this project**:
  `font_import()` only scans system font directories and hard-errors
  (not just "finds nothing") when asked about a font that lives anywhere
  else - exactly the local-file case this project actually needs.
- **Root-caused the warning from `plots-mre.qmd`**: `font family not
  found in Windows font database` is the *base* `png()` device's own
  legacy Windows GDI font lookup failing - confirmed by reproducing it
  with `png()` and watching it vanish with `ragg::agg_png()`. Fixed
  project-wide: `manuscript/_quarto.yml` now sets `knitr: opts_chunk:
  dev: "ragg_png"`.
:::

# Why this notebook exists

[plots-mre.qmd](plots-mre.qmd) hit a real warning
(`font family not found in Windows font database`) while combining a
`flextable` grob with a plot. That warning turned out to be a small
symptom of a much bigger question: fonts in ggplot2 figures are resolved
by a completely different mechanism than the document-level fonts
CLAUDE.md already covers, several competing R packages exist to manage
them, each with its own eccentricities, and more than one graphics device
might end up rendering any given figure. Findings below are from direct
testing in this template - every claim here was actually reproduced, not
assumed from a package's own documentation.

# Two font systems, not one

**Document-level fonts** (html `mainfont`/`monofont`, the docx
reference-doc's theme font, typst `mainfont` - see CLAUDE.md) govern
prose: headings, body text, captions. Pandoc/typst/Word resolve these
themselves, at the document level, once per render.

**Figure-level fonts** are different: a `ggplot2` figure is rendered to a
PNG/SVG by an R *graphics device* before pandoc ever sees it, and that
device has to resolve any font name used inside the plot (via `theme()`,
`element_text(family = ...)`, etc.) on its own. A plot's embedded text has
already been turned into pixels or vector paths by the time pandoc
touches the document - nothing about `mainfont`/`monofont` reaches inside
a figure. This is why "the document renders in the right font, but the
plot's own title doesn't" is a completely different problem from anything
already documented, and needs its own fix.

# The reproducibility gap hiding in plain sight

```{r}
#| label: check-installed-fonts
library(systemfonts)
fonts <- system_fonts()
cat("Fira Sans installed on this machine:", any(grepl("Fira Sans", fonts$family)), "\n")
cat("Fira Code installed on this machine:", any(grepl("Fira Code", fonts$family)), "\n")
```

Both are installed here - confirmed above, not assumed - which is exactly
why using `family = "Fira Sans"` inside a `ggplot2` theme has never
visibly broken during this project's local development. But nothing in
`runme.R`, `_targets.R`, or `.github/workflows/render.yml` installs these
fonts anywhere. A fresh Ubuntu CI runner or a new contributor's machine
has no reason to have them either - the only reason this hasn't already
surfaced as a visible bug is that no figure in this project's actual
manuscript currently sets an explicit font family. The moment one does,
whether it renders correctly becomes a matter of luck about whose machine
it's rendered on, not something this repository guarantees.

# Registering a local font file: systemfonts + ragg

The realistic scenario for a future brand font isn't "add a Google Fonts
entry" - it's "someone hands you a licensed `.otf`/`.ttf` file." A fixture
font (`manuscript/fonts/SpaceMono-Regular.ttf`, generated once by
[`generate-fonts.R`](../generate-fonts.R) and committed to git, the same
pattern `generate-images.R` uses) stands in for that case here - chosen
because it's visually distinctive (a geometric monospace) and confirmed
**not** preinstalled on this machine, so a correct render actually proves
registration worked rather than accidentally matching a font that was
there anyway.

```{r}
#| label: fig-local-font-file
#| fig-cap: "A font registered from an arbitrary local file, not a system font"
register_font(
  name = "Space Mono Local",
  plain = here("manuscript/fonts/SpaceMono-Regular.ttf")
)

ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(title = "Space Mono, registered from a local file") +
  theme_minimal(base_family = "Space Mono Local")
```

No system-wide font install, no admin rights, no OS-specific step - just
a file path. This is confirmed to work with `ragg::agg_png()` (this
project's default figure device, see below) as of `systemfonts` 1.3.2 and
`ragg` 1.5.2.

# Remote web fonts: showtext + sysfonts::font_add_google()

`sysfonts::font_add_google()` fetches a font directly from Google's live
API - genuinely useful for "I just want a Google Font, no file to manage
myself," but it's a real network dependency at render time, unlike the
committed fixture file above. `showtext_auto()` then makes `showtext`
intercept all subsequent text drawing for the rest of this session -
turned back off further down once the point's made.

```{r}
#| label: fig-showtext-google-font
#| fig-cap: "A live Google Font (Space Mono again, fetched remotely this time) via showtext"
library(sysfonts)
library(showtext)

font_add_google("Space Mono", "spacemono_remote")
showtext_auto()

p_showtext <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(title = "Space Mono, fetched live via font_add_google()") +
  theme_minimal(base_family = "spacemono_remote", base_size = 14)
p_showtext
```

## The bug: showtext's DPI has to match the device's, or text is the wrong size

`showtext_opts()` defaults to 96 DPI regardless of what DPI the actual
graphics device renders at. This project's figures render at 300 DPI
(`fig-dpi: 300` in `manuscript/_quarto.yml`) - if `showtext_opts(dpi = )`
is never told that, `showtext`'s text comes out sized for a 96 DPI device
squeezed into a 300 DPI canvas, visibly too small relative to every other
plot element:

```{r}
#| label: fig-showtext-dpi-mismatch
#| fig-cap: "showtext at the default 96 DPI, rendered on a 300 DPI device - text is too small"
#| fig-dpi: 300
p_showtext
```

```{r}
#| label: fig-showtext-dpi-fixed
#| fig-cap: "Same plot, showtext_opts(dpi = 300) matching the actual device - correct size"
#| fig-dpi: 300
showtext_opts(dpi = 300)
p_showtext
```

Confirmed side by side: the fix isn't in the plot code at all, it's
`showtext_opts(dpi = )` matching whatever DPI the device actually renders
at - easy to get wrong once, and easy to forget applies per-document, not
globally, if a project renders at more than one DPI.

```{r}
#| label: showtext-cleanup
#| include: false
# Turn showtext back off - it intercepts every text-drawing call for the
# rest of this R session otherwise, which would affect every chunk below.
showtext_auto(FALSE)
```

# svglite handles registered fonts differently

`svglite` is a vector device (already used elsewhere in this project for
`sample-diagram.svg`, see [images-mre.qmd](images-mre.qmd)) - worth
checking separately from `ragg`'s raster output, since vector and raster
devices don't have to resolve fonts the same way.

```{r}
#| label: check-svglite-font-reference
svglite_path <- tempfile(fileext = ".svg")
p_local <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(title = "Space Mono Local, drawn by svglite") +
  theme_minimal(base_family = "Space Mono Local")
svglite::svglite(svglite_path, width = 8, height = 5)
print(p_local)
dev.off()

svg_lines <- readLines(svglite_path, warn = FALSE)
font_refs <- unique(regmatches(
  svg_lines,
  regexpr('font-family: "[^"]*"', svg_lines)
))
font_refs <- font_refs[nzchar(font_refs)]
cat("What the SVG actually references:", font_refs[1], "\n")
```

Confirms a real difference from `ragg`, not a rendering artifact:
`svglite` writes `font-family: "Space Mono"` - the font's own internal
family name, from the file's metadata - not `"Space Mono Local"`, the
alias `register_font()` was given it under. More consequential than the
naming mismatch itself: the SVG only *references* a font by name, it
doesn't embed the glyphs the way a raster PNG bakes them into pixels.
Opening this SVG on a machine without "Space Mono" (the real name)
available falls back to whatever the browser/viewer picks instead - a
`ragg`-rendered PNG has no equivalent failure mode, since the actual
pixels are already drawn.

# extrafont: why it doesn't fit this project

`extrafont::font_import()` is built around scanning *system* font
directories (`C:\Windows\Fonts` and similar) - not an arbitrary path.
Pointed at the same local file used successfully above:

```{r}
#| label: extrafont-failure
#| error: true
library(extrafont)
font_import(paths = here("manuscript/fonts"), prompt = FALSE)
```

That's not "imports nothing gracefully" - it's a hard error, confirmed
directly, because `extrafont`'s internal code doesn't expect a pattern
match to come back empty. To use `extrafont` at all here, the font would
first need a full, separate, OS-level font installation step outside R
entirely - real friction `systemfonts::register_font()` simply doesn't
have, and irrelevant to how this project's figures actually render
(`ragg`/`svglite`, not the base `pdf()`/`postscript()` devices
`extrafont` was really built around).

# Root-causing (and fixing) the gen_grob() warning

The warning that started this notebook, reproduced directly by forcing
the *base* `png()` device instead of this project's default:

```{r}
#| label: fig-default-device-warning
#| fig-cap: "flextable::gen_grob() output via the base png() device - produces the warning"
#| dev: "png"
#| warning: true
library(flextable)
library(patchwork)

ft <- flextable(head(mtcars[, c("mpg", "wt", "cyl")], 5)) |>
  theme_vanilla() |>
  flextable::bold(part = "header") |>
  autofit()
tbl_grob <- flextable::gen_grob(ft, fit = "fixed", scaling = "min")
p_a <- ggplot(mtcars, aes(wt, mpg)) + geom_point() + theme_minimal()

p_a + wrap_elements(full = tbl_grob)
```

```{r}
#| label: fig-ragg-device-no-warning
#| fig-cap: "Identical figure, this project's actual default device (ragg_png) - no warning"
#| warning: true
p_a + wrap_elements(full = tbl_grob)
```

Confirmed: the warning is specific to the base `png()` device, which on
Windows resolves fonts through the legacy GDI font database (the exact
mechanism named in the warning text itself) - `ragg::agg_png()`, which is
`systemfonts`-aware, resolves the identical font with nothing printed.
**Fixed project-wide, not just documented**: `manuscript/_quarto.yml` now
sets `knitr: opts_chunk: dev: "ragg_png"`, so every figure in this
project - not just this notebook's own chunks - renders through the
device confirmed to handle fonts correctly, including the local-file
registration case above that this project would actually need for a
future brand font.

# Other web font providers

Not tested directly, but worth knowing about if brand-matching comes up
later:

- **Bunny Fonts** - a drop-in mirror of the same Google Fonts catalog,
  EU-hosted, no Google tracking. Same integration pattern as Google Fonts
  (self-host the files, or point at their CDN) - worth it specifically
  for GDPR-conscious institutional use, not because the fonts differ.
- **Adobe Fonts (Typekit)** - a real caveat, not just an alternative:
  licensing normally ties it to Creative Cloud desktop sync, not
  downloadable static files, so it typically doesn't fit a reproducible
  pipeline like this one at all unless the specific font is separately
  licensed as a static file.

# References
