---
# Render via tar_make(), not quarto render directly - see CLAUDE.md.
title: "Maps: sf, geodata sources, projections, and marine layers"
---

```{r}
#| label: setup
#| include: false
library(sf)
library(ggplot2)
library(rnaturalearth)
library(giscoR)
library(eurostat)
library(ggspatial)
library(terra)
library(tidyterra)
library(classInt)
library(shadowtext)
library(patchwork)
library(cowplot)
library(mregions2)
library(marmap)
library(here)
here::i_am("manuscript/supplementary/maps-mre.qmd")

source(here("R/functions.R"))
set_default_palettes()
```

{{< include _back-to-manuscript.qmd >}}

::: {.callout-tip}
# The Short Version
- `sf` + `geom_sf()` + `ggspatial` (scale bar/north arrow) is the default
  drawing stack; reproject to **EPSG:3035** first - an unprojected scale
  bar warns it may be inaccurate.
- `rnaturalearth`/`giscoR`/`eurostat` hand back `sf` objects directly (no
  shapefile downloads); `rnaturalearth`'s `populated_places` and `giscoR`'s
  Urban Audit cover "where are the cities" without ad hoc geocoding.
- `classInt` breaks + this project's viridis/scico defaults is the right
  choropleth combo; `geom_sf_shadowtext()` (`R/functions.R`) keeps labels
  legible over complex fills.
- `patchwork::inset_element()` and `cowplot::draw_plot()` are visually
  interchangeable for insets - but `theme_void()` strips
  `panel.background` (give an inset an explicit opaque fill) and an
  opaque inset hides whatever's beneath it.
- `mregions2` gives IHO Sea Areas/EEZ boundaries directly as `sf`
  (genuinely different shapes, not interchangeable); `marmap` + `terra`
  pulls real NOAA/GEBCO bathymetry - watch `tidyterra`'s `_bathy` palettes,
  which assume every cell is a negative depth.
:::

# Why this notebook exists

This project's manuscripts and figures already lean on `ggplot2`
throughout - maps are not a different tool so much as a different
geometry (`geom_sf()`) and a different category of data source
(boundaries and geodata, not a data frame typed in by hand). This
notebook confirms, by actually fetching real boundaries and rendering
real figures, which R-package-native sources cover this project's actual
needs (Europe generally, marine/water science specifically), what
projection to default to, and two things easy to get wrong the first
time: legible labels over a complex polygon fill, and insetting a
territory that's geographically far from its parent map (Svalbard, for
Norway).

# Drawing: `sf` + `geom_sf()`, and what `ggspatial` adds

`ggplot2::geom_sf()` draws any `sf` object directly - no coordinate
un-nesting, no `fortify()` - and `coord_sf()` handles reprojection at
draw time.

```{r}
#| label: fig-basic-sf
#| fig-cap: "Europe, drawn with plain geom_sf() - no extra packages needed to get this far"
#| fig-width: 7
#| fig-height: 6
europe <- ne_countries(continent = "europe", scale = "medium", returnclass = "sf")

ggplot(europe) +
  geom_sf(fill = "grey85", colour = "grey40") +
  coord_sf(xlim = c(-15, 40), ylim = c(34, 72)) +
  theme_minimal() +
  labs(title = "Europe (rnaturalearth, medium scale)")
```

`geom_sf()` has no built-in scale bar or north arrow - `ggspatial`
supplies both as ordinary layers. Norway is a good stress test for this:
it spans about 13 degrees of latitude, and a `ggspatial` scale bar
computed on that extent while still in geographic (WGS84) coordinates
produces a real, confirmed warning:

```{r}
#| label: fig-ggspatial-unprojected
#| fig-cap: "ggspatial scale bar on an un-projected map - a real warning, not hypothetical"
#| fig-width: 6
#| fig-height: 7
norway_4326 <- ne_countries(country = "Norway", scale = "medium", returnclass = "sf")

ggplot(norway_4326) +
  geom_sf(fill = "grey85") +
  annotation_scale(location = "bl", width_hint = 0.3) +
  theme_minimal() +
  labs(title = "Norway, left in WGS84 - scale bar warning below")
```

```{r}
#| label: fig-ggspatial-projected
#| fig-cap: "Same map, reprojected to EPSG:3035 first - warning gone, scale bar now accurate"
#| fig-width: 6
#| fig-height: 7
norway <- st_transform(norway_4326, 3035)

ggplot(norway) +
  geom_sf(fill = "grey85") +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(
    location = "tr",
    which_north = "true",
    style = north_arrow_fancy_orienteering()
  ) +
  theme_minimal() +
  labs(title = "Norway, projected to EPSG:3035, with scale bar + north arrow")
```

The first chunk prints `Scale on map varies by more than 10%, scale bar
may be inaccurate` - confirmed this is specific to the un-projected CRS,
not to `ggspatial` or this particular map in general: the second chunk,
identical except for the `st_transform(norway_4326, 3035)` reprojection,
produces no warning at all. `annotation_scale()` only produces an
accurate scale bar once the map is on a projected CRS, not left in
geographic lat/lon.

# Boundary and geodata sources, all as R packages

Three packages cover this project's actual needs without touching a
manually-downloaded shapefile:

- **`rnaturalearth`** (+ `rnaturalearthdata`) - general-purpose country/
  state boundaries at three scales (`"small"`/`"medium"`/`"large"`),
  `sf` directly via `ne_countries()`/`ne_states()`. Every example up to
  this point in the notebook uses `scale = "medium"` (`rnaturalearthdata`,
  on CRAN) - `scale = "large"` needs the companion `rnaturalearthhires`
  package too, which is **not** on CRAN (~76MB of high-res shapefiles,
  kept off CRAN and distributed via ropensci's r-universe instead).
  **Now a real dependency, not a hypothetical**: the combined template
  in "Putting it together" below is the first thing in this notebook
  that actually needs `scale = "large"` - confirmed by direct testing
  that `ne_countries(country = "Norway")` returns Bouvet Island
  (~54°S, the notebook's most remote territory) only at `scale =
  "large"`; `scale = "medium"` still includes Jan Mayen but silently
  drops Bouvet Island, and `scale = "small"` drops Jan Mayen too. See
  that section for the rest of this finding (all four territories turn
  out to be one single multipolygon feature, not separate rows).
  `rnaturalearthhires` is accordingly back in `R/required_packages.R`,
  with a matching `extra-repositories:` entry added to the CI
  workflow's `setup-r@v2` step (plain `install.packages()`-style CRAN
  listing can't reach an off-CRAN package regardless of what's in that
  vector).
- **`giscoR`** - Eurostat/GISCO's own boundaries, and **the best source
  specifically for Europe** because it gives NUTS-level regions
  natively, at the actual resolution Eurostat itself publishes.
- **`eurostat`** - pairs with `giscoR`: pulls the statistical data to
  join onto those same NUTS geometries.

```{r}
#| label: fig-gisco-nuts
#| fig-cap: "NUTS2 regions for Norway, via giscoR - includes Svalbard og Jan Mayen"
#| fig-width: 7
#| fig-height: 6
nuts2_no <- gisco_get_nuts(nuts_level = 2, country = "Norway", resolution = "10")

ggplot(nuts2_no) +
  geom_sf(aes(fill = NUTS_NAME)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Norway's NUTS2 regions (giscoR)")
```

`giscoR` returned **seven** NUTS2 regions for Norway, not six - the
seventh is `Svalbard og Jan Mayen`, geographically ~800km north of the
mainland. That single fact is what motivates the inset section below:
any bounding box that includes Svalbard forces the mainland into a
small fraction of the plot area.

# Cities: point data, not boundaries

Boundaries answer "where does a region start and stop" - a different
question from "where are the cities," which is a point-location problem
with its own curation issue: a full gazetteer has hundreds of thousands
of named places, and drawing all of them is never the right answer. The
real question is which subset earns a dot and a label at a given map
extent - the same problem [colour-mre.qmd](colour-mre.qmd) framed for
colour ("not every figure needs a full categorical palette") shows up
here as "not every settlement needs a marker."

`rnaturalearth::ne_download(type = "populated_places")` is the
R-native, general-purpose answer - a curated global point dataset (not
a raw gazetteer dump) that ships exactly the fields needed to make that
call:

```{r}
#| label: fig-cities
#| fig-cap: "Norwegian cities filtered by population, via rnaturalearth's populated_places"
#| fig-width: 6
#| fig-height: 7
places <- ne_download(
  scale = "medium",
  type = "populated_places",
  category = "cultural",
  returnclass = "sf"
)
norway_cities <- places[places$ADM0NAME == "Norway" & places$POP_MAX > 20000, ]

ggplot() +
  geom_sf(data = norway_4326, fill = "grey90") +
  geom_sf(data = norway_cities, colour = "firebrick", size = 1.8) +
  geom_sf_shadowtext(
    data = norway_cities,
    aes(label = NAME),
    bg.colour = "white",
    colour = "black",
    size = 3.5,
    fontface = "bold",
    nudge_x = 1.1,
    hjust = 0
  ) +
  coord_sf(xlim = c(4, 34), ylim = c(57, 71.5)) +
  theme_minimal() +
  labs(title = "Norwegian cities with population > 20,000")
```

The filter above (`POP_MAX > 20000`) is a plain threshold on a field the
dataset already provides - no separate geocoding step, no manual "which
cities matter" list. `SCALERANK`/`LABELRANK` (also included) serve the
same purpose from the cartographic side: lower ranks are the
places Natural Earth itself judges prominent enough to keep at a small
map scale, useful if population isn't the criterion that matters for a
given figure (administrative capitals, say, via the `ADM0CAP`/`ADM1CAP`
flags also present).

**`giscoR::gisco_get_urban_audit()`** is the Europe-specific complement,
returning `sf` points (`spatialtype = "LB"`) for exactly the cities
Eurostat's own Urban Audit programme profiles - a different kind of
curation from `POP_MAX` (statistical-programme membership, not a
population threshold), but ties into the same `giscoR` calls this
notebook already uses for NUTS boundaries:

```{r}
#| label: gisco-urban-audit
ua_cities <- gisco_get_urban_audit(country = "Norway", spatialtype = "LB", level = "CITIES")
sf::st_drop_geometry(ua_cities)[, c("URAU_NAME", "CITY_CPTL")]
```

Only **4** Norwegian cities are in Eurostat's Urban Audit programme
(Oslo, Bergen, Trondheim, Stavanger) against **5** in
`populated_places` above at the `POP_MAX > 20000` threshold (it also
includes Tromsø) - confirms these are two genuinely different curation
criteria, not two sources converging on the same list, so which one to
reach for depends on whether the figure's actual point is population
size or EU/EFTA statistical-programme membership.

For an exhaustive (uncurated) list of settlements rather than a
prominence-filtered one - every town down to a few hundred people, not
just the ones worth a map label - base R's own `maps::world.cities` data
frame is a lighter-weight fallback (`name`/`country.etc`/`pop`/`lat`/
`long`/`capital` columns, `st_as_sf(coords = c("long", "lat"))` away
from an `sf` point layer), at the cost of losing the scale-rank
filtering fields that make `populated_places` the better default for
"which cities belong on this map."

# Projections, with a focus on Europe

Boundaries from all three sources above arrive in **EPSG:4326** (WGS84
lat/lon) - fine for storage, not for area comparison or accurate
distance/scale measurement at European latitudes. **ETRS89-LAEA Europe
(EPSG:3035)** is this notebook's recommended default: it's the
equal-area projection Eurostat itself uses for its own European
thematic maps, and `giscoR`/`eurostat` data lines up with it naturally.

```{r}
#| label: fig-projection-compare
#| fig-cap: "Same NUTS2 data, WGS84 (left) vs EPSG:3035 (right)"
#| fig-width: 9
#| fig-height: 5
nuts2_nordic <- gisco_get_nuts(
  nuts_level = 2,
  country = c("Norway", "Sweden", "Finland"),
  resolution = "20"
)
nuts2_nordic <- nuts2_nordic[!grepl("Svalbard", nuts2_nordic$NUTS_NAME), ]

p_4326 <- ggplot(nuts2_nordic) +
  geom_sf(fill = "grey85") +
  theme_minimal() +
  labs(title = "EPSG:4326 (as supplied)")

p_3035 <- ggplot(st_transform(nuts2_nordic, 3035)) +
  geom_sf(fill = "grey85") +
  theme_minimal() +
  labs(title = "EPSG:3035 (ETRS89-LAEA Europe)")

p_4326 | p_3035
```

The shape difference is subtle at this extent but real - it compounds at
full-Europe extent, and it's what an accurate scale bar depends on (see
`fig-ggspatial` above): `st_transform(x, 3035)` before any figure that
needs to compare area or report distance, not just before adding a
`ggspatial` layer specifically.

**Reserve Web Mercator (EPSG:3857) for compositing with slippy-map
tiles** (`maptiles`, `ggspatial::annotation_map_tile()`) - it's not
appropriate for an area-comparison choropleth, since it distorts area
increasingly with latitude (dramatically so by the time it reaches
Scandinavia).

# Styling: `classInt` breaks + this project's existing palette defaults

[colour-mre.qmd](colour-mre.qmd) found that a raw continuous fill scale
undersells structure that a properly-binned scale reveals - the same
finding holds for a choropleth. `classInt::classIntervals()` computes
real breaks (Jenks natural-breaks, quantile, etc.) instead of leaving
`cut()` to guess even intervals.

```{r}
#| label: fig-choropleth
#| fig-cap: "NUTS2 choropleth: classInt Jenks breaks + this project's viridis default"
#| fig-width: 7
#| fig-height: 7
set.seed(42)
nuts2_3035 <- st_transform(nuts2_nordic, 3035)
nuts2_3035$demo_value <- rlnorm(nrow(nuts2_3035), 3, 0.8)

brks <- classIntervals(nuts2_3035$demo_value, n = 5, style = "jenks")$brks
nuts2_3035$bin <- cut(nuts2_3035$demo_value, brks, include.lowest = TRUE)

ggplot(nuts2_3035) +
  geom_sf(aes(fill = bin), colour = "white", linewidth = 0.3) +
  theme_minimal() +
  labs(
    title = "NUTS2 choropleth, Jenks-binned",
    fill = "Value\n(Jenks bins)"
  )
```

No `scale_fill_*()` is set explicitly here - the fill comes from this
project's own `set_default_palettes()` (`R/functions.R`,
`ggplot2.discrete.fill = "viridis"`), confirmed still applying correctly
to a discrete `sf` fill exactly as it does to any other discrete
`ggplot2` scale.

A base `theme_minimal()`/`theme_void()` with the graticule stripped
(`panel.grid = element_blank()`) is the usual choropleth base - left in
for these figures since the lat/lon graticule is informative context in
a projection-comparison notebook, but for a manuscript figure the
cleaner starting point is `theme_void()`.

# Labels: `geom_sf_shadowtext()` for legibility over complex fills

A plain `geom_sf_text()`/`geom_sf_label()` label can disappear against a
busy multi-coloured background. `geom_sf_shadowtext()` (`R/functions.R`)
gives every label a white halo, the same technique
[GuangchuangYu/shadowtext](https://github.com/GuangchuangYu/shadowtext)
uses for plain `x`/`y` text - ported here to work with `sf` geometry
columns via `stat = "sf_coordinates"`, the same mechanism
`geom_sf_text()` itself uses.

```{r}
#| label: fig-shadowtext
#| fig-cap: "Country labels over a filled polygon layer, with and without a shadow"
#| fig-width: 9
#| fig-height: 5
nordics <- ne_countries(
  country = c("Norway", "Sweden", "Finland", "Denmark"),
  scale = "medium",
  returnclass = "sf"
)
nordics$centroid <- st_centroid(nordics$geometry)

p_plain <- ggplot(nordics) +
  geom_sf(aes(fill = name)) +
  geom_sf_text(aes(geometry = centroid, label = name), colour = "black", size = 4) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "geom_sf_text() - no shadow")

p_shadow <- ggplot(nordics) +
  geom_sf(aes(fill = name)) +
  geom_sf_shadowtext(
    aes(geometry = centroid, label = name),
    bg.colour = "white",
    colour = "black",
    size = 4
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "geom_sf_shadowtext() - white halo")

p_plain | p_shadow
```

Both labels sit on the same fill colours here, so the difference is
subtle at this figure's size - it stops being subtle the moment a label
lands on a genuinely dark or highly-saturated fill (a dark end of a
viridis/scico scale, in particular), which is exactly the case this
technique is for.

# Insets: `patchwork::inset_element()` vs `cowplot::draw_plot()`

Norway's NUTS2 set above included Svalbard, ~800km north of the
mainland - the standard cartographic answer is an inset, not a bounding
box that includes both and squashes the mainland. **Confirmed
`patchwork` alone can do this** - `cowplot` is not required just for
this technique, contrary to an easy assumption that `cowplot` is the
only option.

```{r}
#| label: fig-inset-patchwork
#| fig-cap: "Mainland Norway with a Svalbard inset, via patchwork::inset_element()"
#| fig-width: 6
#| fig-height: 7
mainland <- ggplot(norway) +
  geom_sf(fill = "grey85") +
  coord_sf(xlim = c(4, 32), ylim = c(57, 71.5), expand = FALSE, crs = 4326) +
  theme_minimal() +
  labs(title = "Mainland Norway")

svalbard <- ggplot(norway) +
  geom_sf(fill = "grey85") +
  coord_sf(xlim = c(9, 34), ylim = c(74, 81), expand = FALSE, crs = 4326) +
  theme_void() +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  )

mainland + inset_element(svalbard, left = 0.55, bottom = 0.65, right = 1, top = 1)
```

```{r}
#| label: fig-inset-cowplot
#| fig-cap: "The same inset, via cowplot::draw_plot() - visually equivalent"
#| fig-width: 6
#| fig-height: 7
ggdraw(mainland) +
  draw_plot(svalbard, x = 0.55, y = 0.65, width = 0.42, height = 0.32)
```

Two things confirmed while testing this, both easy to miss:

- **`theme_void()` sets `panel.background = element_blank()`** - fully
  transparent, not just borderless. An inset built with a bare
  `theme_void()` renders with no visible fill at all against a white
  page, indistinguishable from "the inset didn't render" unless an
  explicit `panel.background = element_rect(fill = "white")` (or
  similar) is added, as above.
- **Loading both packages in one session masks
  `patchwork::align_plots()` with `cowplot`'s own version** (`cowplot`
  attached after `patchwork`, per this notebook's own setup chunk) -
  confirmed via R's own startup message, not assumed. Only relevant if
  a document actually calls `align_plots()` - `inset_element()` itself
  is unaffected - but worth knowing before reaching for that specific
  function with both packages loaded.

**Recommendation: default to `patchwork::inset_element()`**, not
`cowplot`, given this project already depends on `patchwork` for every
other plot-combining need (see [plots-mre.qmd](plots-mre.qmd)) - adding
`cowplot` only pays for itself if a specific pixel-precise placement
need comes up that `inset_element()`'s `align_to` options don't cover.

# Marine/ocean data: `mregions2` and `marmap`

General-purpose land boundaries (`rnaturalearth`/`giscoR`) have no
concept of a named sea area or an Exclusive Economic Zone - that's a
different data domain, maintained by the Flanders Marine Institute
(VLIZ) and published at marineregions.org, including the IHO's own
"Limits of Oceans and Seas" boundaries. **`mregions2` is the R-native
access point for all of it**, querying VLIZ's OGC API directly and
returning `sf`.

```{r}
#| label: fig-iho-vs-eez
#| fig-cap: "IHO Sea Areas vs Norway's actual EEZ boundary - genuinely different shapes"
#| fig-width: 8
#| fig-height: 8
iho_north <- mrp_get(
  "iho",
  cql_filter = "name IN ('Norwegian Sea', 'North Sea')"
)
eez_norway <- mrp_get("eez", cql_filter = "territory1 = 'Norway'")

ggplot() +
  geom_sf(data = iho_north, aes(fill = name), alpha = 0.4, colour = NA) +
  geom_sf(data = eez_norway, fill = NA, colour = "black", linewidth = 0.6, linetype = "dashed") +
  geom_sf(data = europe, fill = "grey90", colour = "grey40") +
  coord_sf(xlim = c(-15, 40), ylim = c(55, 82)) +
  scale_fill_brewer(palette = "Set2", name = "IHO Sea Area") +
  theme_minimal() +
  labs(
    title = "IHO Sea Areas (fill) vs Norway's EEZ (dashed outline)",
    subtitle = "A legal/economic boundary and a 1953 cartographic convention - not the same shape"
  )
```

The IHO fill and the EEZ outline visibly diverge - confirms these are
two genuinely different boundary concepts, not two sources for the same
thing. **Two gotchas found while building that figure**:

- **`mregions2::mrp_list` is a static data frame, not a function** -
  despite the naming convention of everything else in the package
  (`mrp_get()`, `mrp_view()`, ...), calling it as `mrp_list()` fails with
  `could not find function "mrp_list"`. Reference it bare (`mrp_list`)
  to browse the available layer catalogue and find the right `layer`
  code (`"iho"`, `"eez"`, ...) for `mrp_get()`.
- **IHO names don't always match the expected English spelling** -
  querying for `'%Barents%'` returns nothing; the actual name in this
  gazetteer is `"Barentsz Sea"` (the historical Dutch spelling, after
  Willem Barentsz). Use `gaz_search()` or a wildcard `LIKE` filter to
  confirm a name before assuming a `cql_filter` returned everything it
  should have.

Bathymetry is a separate concern from named-area boundaries -
**`marmap`** wraps a live NOAA/GEBCO grid fetch, and converts cleanly
into this same `sf`/`ggplot2` pipeline via `terra`/`tidyterra`:

```{r}
#| label: fig-bathymetry
#| fig-cap: "GEBCO bathymetry off the Norwegian coast, via marmap + tidyterra"
#| fig-width: 7
#| fig-height: 6
bathy <- getNOAA.bathy(
  lon1 = 0,
  lon2 = 15,
  lat1 = 58,
  lat2 = 66,
  resolution = 10,
  keep = TRUE
)
bathy_rast <- terra::rast(marmap::as.raster(bathy))

ggplot() +
  geom_spatraster(data = bathy_rast) +
  scale_fill_hypso_tint_c(
    palette = "colombia_bathy",
    guide = guide_colorbar(title = "Depth (m)")
  ) +
  theme_minimal() +
  labs(title = "Bathymetry, Norwegian coastal shelf (GEBCO via marmap)")
```

The white patches are land, not missing data - confirmed by inspecting
the raster's own elevation values there (positive, not `NA`).
`scale_fill_hypso_tint_c()`'s `_bathy`-suffixed palettes assume every
cell is a negative depth; a land cell's positive elevation falls outside
that assumed range and renders blank rather than erroring, which reads
as a data gap unless checked. Fine for a bathymetry-only figure (crop to
water before plotting, or accept blank land as context); reach for a
combined hypsometric-tint palette instead if land relief needs to show
too.

# Putting it together: a Norway map template

Everything above is a separate technique, confirmed in isolation.
Combining them into one figure is a different, harder problem - it's
also the actually-useful end product for a real manuscript (this
project's own domain being Norwegian marine/water science), so this
section builds one: mainland Norway with NUTS3 boundaries, an inset per
remote territory (Svalbard, including Bjørnøya/Bear Island; Jan Mayen;
Bouvet Island), country/territory name labels, city labels, and the rest
of Europe present but visually out of the way.

## Data: one `ne_countries()` call covers every territory

```{r}
#| label: template-data
norway_hires <- ne_countries(country = "Norway", scale = "large", returnclass = "sf")
europe_context <- ne_countries(continent = "Europe", scale = "medium", returnclass = "sf")
europe_context <- europe_context[europe_context$name != "Norway", ]

nuts3_no <- gisco_get_nuts(nuts_level = 3, country = "Norway", resolution = "10")
nuts3_mainland <- nuts3_no[!nuts3_no$NUTS_ID %in% c("NO0B1", "NO0B2"), ]

# norway_cities/places already fetched in "Cities: point data, not
# boundaries" above - same chunk-execution session, no need to re-fetch.
svalbard_places <- places[places$NAME == "Longyearbyen", ]

country_labels <- st_as_sf(
  data.frame(
    label = c("NORWAY", "Sweden", "Finland"),
    lon = c(9.5, 15, 22),
    lat = c(61.2, 62, 68)
  ),
  coords = c("lon", "lat"), crs = 4326
)

bear_island_label <- st_as_sf(
  data.frame(label = "Bjørnøya\n(Bear Island)", lon = 19.0, lat = 74.35),
  coords = c("lon", "lat"), crs = 4326
)
```

**`ne_countries(country = "Norway", scale = "large")` returns a single
row** - one `MULTIPOLYGON` geometry that already bundles mainland Norway,
Svalbard (Bjørnøya/Bear Island included, as one of its ~120 constituent
polygon parts), Jan Mayen, *and* Bouvet Island together under the one
`name == "Norway"` feature. Confirmed directly: none of these four are
separate named rows at any scale - searching `ne_countries(scale =
...)`'s `name`/`name_long` columns for `"Svalbard"`, `"Jan Mayen"`, or
`"Bouvet"` returns zero matches regardless of scale, because they were
never separate features to find. The only way to isolate one territory
from another is to crop by its own bounding box (`st_crop()`, or
`coord_sf(xlim=, ylim=)` at draw time, as the inset builder below does),
not to filter by name.

**Confirmed by direct testing which `scale` actually keeps which
territory** - this is the concrete reason this section needs `scale =
"large"` when nothing earlier in the notebook did:

| `scale` | mainland | Svalbard | Jan Mayen | Bouvet Island |
|---|---|---|---|---|
| `"large"`  | yes | yes | yes | **yes** |
| `"medium"` | yes | yes | yes | no |
| `"small"`  | yes | yes | no  | no |

`scale = "medium"` (`rnaturalearthdata`, on CRAN, what every earlier
example in this notebook uses) silently drops Bouvet Island from the
same bundled multipolygon rather than erroring - confirmed by cropping
each scale's returned geometry to Bouvet's known bounding box (~54.4°S,
3.4°E) and checking whether anything survives. Only `scale = "large"`
(`rnaturalearthhires`, off-CRAN - see "Boundary and geodata sources"
above for the CI implication) keeps all four.

## The mainland panel

```{r}
#| label: template-mainland
mainland <- ggplot() +
  geom_sf(data = europe_context, fill = "grey92", colour = "grey80", linewidth = 0.15) +
  geom_sf(data = nuts3_mainland, fill = "white", colour = "grey30", linewidth = 0.35) +
  geom_sf_text(
    data = country_labels[country_labels$label %in% c("Sweden", "Finland"), ],
    aes(label = label), colour = "grey40", size = 3, fontface = "italic"
  ) +
  geom_sf_text(
    data = country_labels[country_labels$label == "NORWAY", ],
    aes(label = label), colour = "grey15", size = 4.5, fontface = "bold"
  ) +
  geom_sf(data = norway_cities, colour = "black", size = 1.5) +
  geom_sf_shadowtext(
    data = norway_cities, aes(label = NAME),
    bg.colour = "white", colour = "black", size = 3, fontface = "bold",
    nudge_x = 0.6, hjust = 0
  ) +
  coord_sf(xlim = c(3, 32), ylim = c(57.5, 71.5), expand = FALSE, crs = 4326) +
  theme_bw() +
  theme(
    panel.grid = element_line(colour = "grey85", linewidth = 0.25),
    axis.text = element_text(size = 7, colour = "grey30"),
    axis.ticks = element_line(colour = "grey50", linewidth = 0.3),
    axis.title = element_blank()
  ) +
  labs(title = "Norway: NUTS3 regions", caption = "Boundaries: GISCO/Eurostat. Basemap: Natural Earth.")
```

This is the split the brief asked for: **`geom_sf_text()` (plain, no
halo) for country/territory names** - Norway itself and its neighbours,
sparse enough that a plain label reads fine against the de-emphasised
backdrop - **vs. `geom_sf_shadowtext()` for cities**, which sit directly
on top of Norway's own NUTS3 fill and need the halo for the same reason
the "Labels" section above demonstrated. "De-emphasise the rest of
Europe" is deliberately not just a fainter fill: `europe_context` gets a
flat pale grey with thin, low-contrast borders (`grey80`, `linewidth =
0.15`), while `nuts3_mainland` gets a plain white fill with visibly
darker, thicker boundaries (`grey30`, `linewidth = 0.35`) - the boundary
weight difference is what actually separates "the subject" from "the
context" at a glance, more than the fill colour does on its own. Norway
is left plain white rather than given an illustrative choropleth fill
(unlike `fig-choropleth` above) precisely because this is meant as a
*template* - a real manuscript figure drops its own `aes(fill = ...)`
onto `nuts3_mainland` here.

`theme_bw()`, not `theme_minimal()`, is the base theme here - deliberately
different from the plain `theme_minimal()` used for most other figures in
this notebook, specifically to keep the graticule (as `panel.grid`,
drawn straight rather than curved since this whole figure stays in
un-projected EPSG:4326) and the degree-formatted axis text/ticks
`coord_sf()` computes automatically. Both are toned down (`grey85`
gridlines, `grey30` 7pt axis text) rather than left at `theme_bw()`'s
own black-on-white default, so they read as reference context rather
than competing with the labels and boundaries that are the actual point
of the figure.

## One inset builder, three territories

```{r}
#| label: template-insets
build_inset <- function(xlim, ylim, title) {
  ggplot() +
    geom_sf(data = norway_hires, fill = "grey85", colour = "grey40", linewidth = 0.3) +
    coord_sf(xlim = xlim, ylim = ylim, expand = FALSE, crs = 4326) +
    theme_bw() +
    theme(
      panel.grid = element_line(colour = "grey85", linewidth = 0.2),
      axis.text = element_text(size = 5, colour = "grey30"),
      axis.ticks = element_line(colour = "grey50", linewidth = 0.25),
      axis.title = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),
      plot.title = element_text(size = 8, hjust = 0.5, margin = margin(b = 2))
    ) +
    labs(title = title)
}

svalbard <- build_inset(xlim = c(9, 34), ylim = c(74, 81), title = "Svalbard") +
  geom_sf_text(
    data = bear_island_label, aes(label = label),
    size = 2.2, lineheight = 0.85, nudge_x = 3.5
  )
jan_mayen <- build_inset(xlim = c(-10, -7), ylim = c(70.7, 71.2), title = "Jan Mayen")
bouvet <- build_inset(xlim = c(3, 4), ylim = c(-54.6, -54.3), title = "Bouvet Island")
```

Each inset crops the *same* `norway_hires` object to its own territory's
bounding box. Unlike the single-inset example in the "Insets" section
above, these use `theme_bw()` rather than `theme_void()` - for the same
graticule/axis-text/tick reason as the mainland panel - which
incidentally sidesteps that section's `panel.background` gotcha
entirely: `theme_bw()`'s default panel background is already opaque
white, so nothing here needs the explicit `element_rect(fill = "white")`
override `theme_void()` required. The explicit `panel.border` override
is still needed, though - `theme_bw()`'s own default border is a thin
grey, and a visibly black, thicker frame is what actually reads as "this
is an inset, not part of the base map" at a glance. Bjørnøya gets its
own `geom_sf_text()` label layered onto the Svalbard inset specifically,
since it's a real, separately-named feature a reader would otherwise
have no way to pick out from the rest of the archipelago.

## Composite: three insets, not one

```{r}
#| label: fig-norway-template
#| fig-cap: "A complete Norway map template: mainland NUTS3 + de-emphasised Europe + three remote-territory insets"
#| fig-width: 8
#| fig-height: 10
mainland +
  inset_element(svalbard, left = 0.70, bottom = 0.72, right = 1.0, top = 1.0) +
  inset_element(jan_mayen, left = 0.70, bottom = 0.40, right = 1.0, top = 0.68) +
  inset_element(bouvet, left = 0.70, bottom = 0.02, right = 1.0, top = 0.30)
```

Stacking three `inset_element()` calls instead of one is a mechanical
extension of the pattern already used above - `+` composes any number
of them onto the same base plot - but it surfaced a real bug the
single-inset version never could: **the first version of this figure
placed all three insets in a column spanning most of the panel's right
edge, from `bottom = 0.02` to `top = 1.0`, and that column happened to
overlap the "Finland" label's actual plotted position.** `inset_element()`
draws an opaque panel (needed for the reason above), so it doesn't blend
with or show through whatever the base plot already drew there - it
simply covers it, with no warning that anything was hidden. The label
wasn't misplaced or dropped by the code; it rendered exactly where
specified and was then painted over. Confirmed only by rendering the
figure and looking - the R code for both the label and the insets was
individually correct, and nothing about reading either chunk in
isolation would suggest a conflict between them. Fixed by choosing
inset bounds that stay clear of every other layer's actual plotted
extent, not just leaving visual "gaps" in the intended layout - the two
aren't the same thing once labels can sit anywhere in data space.

::: {.callout-note}
## Bouvet Island's inset shows almost nothing, and that's correct

Bouvet Island is roughly 49 km² - smaller than most cities. At the
zoom level needed to make it visible as more than a single pixel, its
inset legitimately shows just one small grey shape on a white
background, no coastline detail, no context. That's the real geography,
not an under-cropped extent.
:::

::: {.callout-note}
## A real, narrow encoding gotcha found while writing this section

`"Bjørnøya"`/`"Tromsø"` typed directly as an R string literal rendered
correctly here (Quarto reads `.qmd` source as UTF-8) - but the same
literal, tried in a plain `Rscript -e "..."` call from a shell **outside
Quarto** while drafting this section, came out corrupted on this
Windows machine. Confirmed the corruption was specific to how that
string reached R, not a font/glyph problem in the `ragg_png` device
itself: `"Tromsø"` (the same character, arriving via `ne_download()`'s
own data rather than a hand-typed literal) rendered correctly in the
very same corrupted image. Not a Quarto/render-pipeline issue - flagged
here because a future edit to this notebook done by testing snippets in
a bare `Rscript` call first, outside `quarto render`/`tar_make()`,
could otherwise mistake a shell-specific artifact for a real bug in the
figure itself.
:::

::: {.callout-note}
## Network dependency, not just a footnote

Every fetch in this section (`rnaturalearth`/`giscoR`'s first download,
`mregions2`'s OGC queries, `marmap`'s NOAA grid request) is a live
network call, made fresh on every render given this project's
`freeze: false` setting (see CLAUDE.md's "Related, separate gotcha" on
why freeze is off). `marmap::getNOAA.bathy(..., keep = TRUE)` caches its
result to a local CSV keyed by the bounding box/resolution, so repeat
renders of the *same* extent reuse it; `rnaturalearth`/`giscoR` cache
downloaded boundary files similarly. A CI runner with no prior cache, or
a network-restricted environment, will hit these servers fresh - same
category of risk already flagged for `cetz`'s package fetch in
[diagrams-mre.qmd](diagrams-mre.qmd), not independently re-verified
against a restricted CI runner here.
:::

# References
