---
# Render via `targets::tar_make()` (the `render_manuscript` target in
# _targets.R), not `quarto render` directly - see manuscript.qmd.
#
# Maps: sf + ggplot2::geom_sf() as the drawing layer, R-package-native
# boundary/statistical sources (rnaturalearth, giscoR, eurostat) instead
# of hand-downloaded shapefiles, curated city-point sources (rnaturalearth
# populated_places, giscoR Urban Audit) so "where are the cities" doesn't
# turn into an ad hoc geocoding project, projections with a Europe focus
# (EPSG:3035), choropleth styling with classInt breaks, geom_sf_shadowtext()
# for labels over complex backgrounds, insets (patchwork vs cowplot), and
# marine-specific data (mregions2's IHO Sea Areas/EEZ, marmap bathymetry) -
# this project's actual domain (NIVA, marine/water science).
title: "Maps: sf, geodata sources, projections, and marine layers"
---

```{r}
#| label: setup
#| include: false
library(sf)
library(ggplot2)
library(rnaturalearth)
library(giscoR)
library(eurostat)
library(ggspatial)
library(terra)
library(tidyterra)
library(classInt)
library(shadowtext)
library(patchwork)
library(cowplot)
library(mregions2)
library(marmap)
library(here)
here::i_am("manuscript/supplementary/maps-mre.qmd")

source(here("R/functions.R"))
set_default_palettes()
```

{{< include _back-to-manuscript.qmd >}}

::: {.callout-tip}
# The Short Version
- **`sf` + `ggplot2::geom_sf()` remains the right default for drawing** -
  it composes with everything else this template already does
  (`theme_set()`, `patchwork`, the viridis/scico defaults from
  [colour-mre.qmd](colour-mre.qmd), `ragg_png`). `ggspatial` adds the one
  thing `geom_sf()` has no native equivalent for: a scale bar and north
  arrow.
- **`rnaturalearth`, `giscoR`, and `eurostat` hand back `sf` objects
  directly** - no manual shapefile download for any of the boundaries
  used in this notebook. `giscoR` is the standout for **Europe**
  specifically: it's Eurostat/GISCO's own NUTS boundaries.
- **"Where are the cities" has an R-native answer, not just a future
  geonorge lookup**: `rnaturalearth::ne_download(type =
  "populated_places")` is a curated global `POINT` dataset that ships
  its own `POP_MAX`/`SCALERANK`/`LABELRANK` fields - confirmed these
  exist specifically to answer "which cities are prominent enough to
  draw at this extent," not just a name/lat/lon dump. `giscoR::
  gisco_get_urban_audit()` is the Europe-specific complement, curated by
  Eurostat's own Urban Audit statistical programme membership rather
  than a population threshold.
- **Reproject before adding a `ggspatial` scale bar - confirmed this
  isn't cosmetic.** An un-projected (WGS84 lat/lon) map of Norway
  triggered a real warning, `Scale on map varies by more than 10%, scale
  bar may be inaccurate` - confirmed it disappears once the same data is
  `st_transform()`'d to **EPSG:3035 (ETRS89-LAEA Europe)** first, this
  notebook's recommended default CRS for any European-extent figure.
- **`classInt::classIntervals()` breaks (Jenks/quantile) + this
  project's existing viridis/scico defaults is the right choropleth
  combo** - same "raw continuous undersells structure" finding
  [colour-mre.qmd](colour-mre.qmd) already made for a non-spatial
  heatmap, confirmed again here for polygon fills.
- **`geom_sf_shadowtext()`** (`R/functions.R`, ported from
  [GuangchuangYu/shadowtext](https://github.com/GuangchuangYu/shadowtext)
  for `sf` geometries) - confirmed working for labelling polygons/points
  where a plain `geom_sf_text()` label would disappear into a
  multi-coloured background.
- **`patchwork::inset_element()` and `cowplot::draw_plot()` produced
  visually identical insets, confirmed side by side** - `cowplot` is not
  required just to inset one map inside another. Two real gotchas found
  while confirming that: `theme_void()` strips `panel.background`
  entirely, so an inset is genuinely transparent (not just borderless)
  unless given an explicit opaque fill; and loading both packages in one
  session masks `patchwork::align_plots()` with `cowplot`'s own version.
- **`mregions2` is the R-native access point for the IHO's own
  boundaries** (via Flanders Marine Institute/marineregions.org, who
  maintain the canonical digital version) - confirmed fetching IHO Sea
  Areas and EEZ layers directly as `sf`. Confirmed concretely that they
  are genuinely different shapes, not interchangeable: Norway's EEZ and
  the IHO's "Norwegian Sea" polygon don't align, because one is a legal/
  economic boundary and the other a 1953 cartographic convention.
- **`marmap` + `terra`/`tidyterra` pulls real NOAA/GEBCO bathymetry into
  the same `sf`/`ggplot2` pipeline** - confirmed with a live fetch. Watch
  `tidyterra`'s `_bathy`-suffixed hypsometric palettes specifically: they
  assume every cell is a negative depth, so land cells in the same
  raster render blank, not an obviously-wrong colour.
:::

# Why this notebook exists

This project's manuscripts and figures already lean on `ggplot2`
throughout - maps are not a different tool so much as a different
geometry (`geom_sf()`) and a different category of data source
(boundaries and geodata, not a data frame typed in by hand). This
notebook confirms, by actually fetching real boundaries and rendering
real figures, which R-package-native sources cover this project's actual
needs (Europe generally, marine/water science specifically), what
projection to default to, and two things easy to get wrong the first
time: legible labels over a complex polygon fill, and insetting a
territory that's geographically far from its parent map (Svalbard, for
Norway).

# Drawing: `sf` + `geom_sf()`, and what `ggspatial` adds

`ggplot2::geom_sf()` draws any `sf` object directly - no coordinate
un-nesting, no `fortify()` - and `coord_sf()` handles reprojection at
draw time.

```{r}
#| label: fig-basic-sf
#| fig-cap: "Europe, drawn with plain geom_sf() - no extra packages needed to get this far"
#| fig-width: 7
#| fig-height: 6
europe <- ne_countries(continent = "europe", scale = "medium", returnclass = "sf")

ggplot(europe) +
  geom_sf(fill = "grey85", colour = "grey40") +
  coord_sf(xlim = c(-15, 40), ylim = c(34, 72)) +
  theme_minimal() +
  labs(title = "Europe (rnaturalearth, medium scale)")
```

`geom_sf()` has no built-in scale bar or north arrow - `ggspatial`
supplies both as ordinary layers. Norway is a good stress test for this:
it spans about 13 degrees of latitude, and a `ggspatial` scale bar
computed on that extent while still in geographic (WGS84) coordinates
produces a real, confirmed warning:

```{r}
#| label: fig-ggspatial-unprojected
#| fig-cap: "ggspatial scale bar on an un-projected map - a real warning, not hypothetical"
#| fig-width: 6
#| fig-height: 7
norway_4326 <- ne_countries(country = "Norway", scale = "medium", returnclass = "sf")

ggplot(norway_4326) +
  geom_sf(fill = "grey85") +
  annotation_scale(location = "bl", width_hint = 0.3) +
  theme_minimal() +
  labs(title = "Norway, left in WGS84 - scale bar warning below")
```

```{r}
#| label: fig-ggspatial-projected
#| fig-cap: "Same map, reprojected to EPSG:3035 first - warning gone, scale bar now accurate"
#| fig-width: 6
#| fig-height: 7
norway <- st_transform(norway_4326, 3035)

ggplot(norway) +
  geom_sf(fill = "grey85") +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(
    location = "tr",
    which_north = "true",
    style = north_arrow_fancy_orienteering()
  ) +
  theme_minimal() +
  labs(title = "Norway, projected to EPSG:3035, with scale bar + north arrow")
```

The first chunk prints `Scale on map varies by more than 10%, scale bar
may be inaccurate` - confirmed this is specific to the un-projected CRS,
not to `ggspatial` or this particular map in general: the second chunk,
identical except for the `st_transform(norway_4326, 3035)` reprojection,
produces no warning at all. `annotation_scale()` only produces an
accurate scale bar once the map is on a projected CRS, not left in
geographic lat/lon.

# Boundary and geodata sources, all as R packages

Three packages cover this project's actual needs without touching a
manually-downloaded shapefile:

- **`rnaturalearth`** (+ `rnaturalearthdata`) - general-purpose country/
  state boundaries at three scales (`"small"`/`"medium"`/`"large"`),
  `sf` directly via `ne_countries()`/`ne_states()`. Every example in
  this notebook uses `scale = "medium"` (`rnaturalearthdata`, on CRAN) -
  `scale = "large"` needs the companion `rnaturalearthhires` package
  too, which is **not** on CRAN (confirmed by a real CI failure: it's
  ~76MB of high-res shapefiles, kept off CRAN and distributed via
  ropensci's r-universe instead) and so isn't in this project's
  `R/required_packages.R` - install it separately
  (`install.packages("rnaturalearthhires", repos =
  "https://ropensci.r-universe.dev")`) only if a future edit actually
  needs `scale = "large"`.
- **`giscoR`** - Eurostat/GISCO's own boundaries, and **the best source
  specifically for Europe** because it gives NUTS-level regions
  natively, at the actual resolution Eurostat itself publishes.
- **`eurostat`** - pairs with `giscoR`: pulls the statistical data to
  join onto those same NUTS geometries.

```{r}
#| label: fig-gisco-nuts
#| fig-cap: "NUTS2 regions for Norway, via giscoR - includes Svalbard og Jan Mayen"
#| fig-width: 7
#| fig-height: 6
nuts2_no <- gisco_get_nuts(nuts_level = 2, country = "Norway", resolution = "10")

ggplot(nuts2_no) +
  geom_sf(aes(fill = NUTS_NAME)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Norway's NUTS2 regions (giscoR)")
```

`giscoR` returned **seven** NUTS2 regions for Norway, not six - the
seventh is `Svalbard og Jan Mayen`, geographically ~800km north of the
mainland. That single fact is what motivates the inset section below:
any bounding box that includes Svalbard forces the mainland into a
small fraction of the plot area.

# Cities: point data, not boundaries

Boundaries answer "where does a region start and stop" - a different
question from "where are the cities," which is a point-location problem
with its own curation issue: a full gazetteer has hundreds of thousands
of named places, and drawing all of them is never the right answer. The
real question is which subset earns a dot and a label at a given map
extent - the same problem [colour-mre.qmd](colour-mre.qmd) framed for
colour ("not every figure needs a full categorical palette") shows up
here as "not every settlement needs a marker."

`rnaturalearth::ne_download(type = "populated_places")` is the
R-native, general-purpose answer - a curated global point dataset (not
a raw gazetteer dump) that ships exactly the fields needed to make that
call:

```{r}
#| label: fig-cities
#| fig-cap: "Norwegian cities filtered by population, via rnaturalearth's populated_places"
#| fig-width: 6
#| fig-height: 7
places <- ne_download(
  scale = "medium",
  type = "populated_places",
  category = "cultural",
  returnclass = "sf"
)
norway_cities <- places[places$ADM0NAME == "Norway" & places$POP_MAX > 20000, ]

ggplot() +
  geom_sf(data = norway_4326, fill = "grey90") +
  geom_sf(data = norway_cities, colour = "firebrick", size = 1.8) +
  geom_sf_shadowtext(
    data = norway_cities,
    aes(label = NAME),
    bg.colour = "white",
    colour = "black",
    size = 3.5,
    fontface = "bold",
    nudge_x = 1.1,
    hjust = 0
  ) +
  coord_sf(xlim = c(4, 34), ylim = c(57, 71.5)) +
  theme_minimal() +
  labs(title = "Norwegian cities with population > 20,000")
```

The filter above (`POP_MAX > 20000`) is a plain threshold on a field the
dataset already provides - no separate geocoding step, no manual "which
cities matter" list. `SCALERANK`/`LABELRANK` (also included) serve the
same purpose from the cartographic side: lower ranks are the
places Natural Earth itself judges prominent enough to keep at a small
map scale, useful if population isn't the criterion that matters for a
given figure (administrative capitals, say, via the `ADM0CAP`/`ADM1CAP`
flags also present).

**`giscoR::gisco_get_urban_audit()`** is the Europe-specific complement,
returning `sf` points (`spatialtype = "LB"`) for exactly the cities
Eurostat's own Urban Audit programme profiles - a different kind of
curation from `POP_MAX` (statistical-programme membership, not a
population threshold), but ties into the same `giscoR` calls this
notebook already uses for NUTS boundaries:

```{r}
#| label: gisco-urban-audit
ua_cities <- gisco_get_urban_audit(country = "Norway", spatialtype = "LB", level = "CITIES")
sf::st_drop_geometry(ua_cities)[, c("URAU_NAME", "CITY_CPTL")]
```

Only **4** Norwegian cities are in Eurostat's Urban Audit programme
(Oslo, Bergen, Trondheim, Stavanger) against **5** in
`populated_places` above at the `POP_MAX > 20000` threshold (it also
includes Tromsø) - confirms these are two genuinely different curation
criteria, not two sources converging on the same list, so which one to
reach for depends on whether the figure's actual point is population
size or EU/EFTA statistical-programme membership.

For an exhaustive (uncurated) list of settlements rather than a
prominence-filtered one - every town down to a few hundred people, not
just the ones worth a map label - base R's own `maps::world.cities` data
frame is a lighter-weight fallback (`name`/`country.etc`/`pop`/`lat`/
`long`/`capital` columns, `st_as_sf(coords = c("long", "lat"))` away
from an `sf` point layer), at the cost of losing the scale-rank
filtering fields that make `populated_places` the better default for
"which cities belong on this map."

# Projections, with a focus on Europe

Boundaries from all three sources above arrive in **EPSG:4326** (WGS84
lat/lon) - fine for storage, not for area comparison or accurate
distance/scale measurement at European latitudes. **ETRS89-LAEA Europe
(EPSG:3035)** is this notebook's recommended default: it's the
equal-area projection Eurostat itself uses for its own European
thematic maps, and `giscoR`/`eurostat` data lines up with it naturally.

```{r}
#| label: fig-projection-compare
#| fig-cap: "Same NUTS2 data, WGS84 (left) vs EPSG:3035 (right)"
#| fig-width: 9
#| fig-height: 5
nuts2_nordic <- gisco_get_nuts(
  nuts_level = 2,
  country = c("Norway", "Sweden", "Finland"),
  resolution = "20"
)
nuts2_nordic <- nuts2_nordic[!grepl("Svalbard", nuts2_nordic$NUTS_NAME), ]

p_4326 <- ggplot(nuts2_nordic) +
  geom_sf(fill = "grey85") +
  theme_minimal() +
  labs(title = "EPSG:4326 (as supplied)")

p_3035 <- ggplot(st_transform(nuts2_nordic, 3035)) +
  geom_sf(fill = "grey85") +
  theme_minimal() +
  labs(title = "EPSG:3035 (ETRS89-LAEA Europe)")

p_4326 | p_3035
```

The shape difference is subtle at this extent but real - it compounds at
full-Europe extent, and it's what an accurate scale bar depends on (see
`fig-ggspatial` above): `st_transform(x, 3035)` before any figure that
needs to compare area or report distance, not just before adding a
`ggspatial` layer specifically.

**Reserve Web Mercator (EPSG:3857) for compositing with slippy-map
tiles** (`maptiles`, `ggspatial::annotation_map_tile()`) - it's not
appropriate for an area-comparison choropleth, since it distorts area
increasingly with latitude (dramatically so by the time it reaches
Scandinavia).

# Styling: `classInt` breaks + this project's existing palette defaults

[colour-mre.qmd](colour-mre.qmd) found that a raw continuous fill scale
undersells structure that a properly-binned scale reveals - the same
finding holds for a choropleth. `classInt::classIntervals()` computes
real breaks (Jenks natural-breaks, quantile, etc.) instead of leaving
`cut()` to guess even intervals.

```{r}
#| label: fig-choropleth
#| fig-cap: "NUTS2 choropleth: classInt Jenks breaks + this project's viridis default"
#| fig-width: 7
#| fig-height: 7
set.seed(42)
nuts2_3035 <- st_transform(nuts2_nordic, 3035)
nuts2_3035$demo_value <- rlnorm(nrow(nuts2_3035), 3, 0.8)

brks <- classIntervals(nuts2_3035$demo_value, n = 5, style = "jenks")$brks
nuts2_3035$bin <- cut(nuts2_3035$demo_value, brks, include.lowest = TRUE)

ggplot(nuts2_3035) +
  geom_sf(aes(fill = bin), colour = "white", linewidth = 0.3) +
  theme_minimal() +
  labs(
    title = "NUTS2 choropleth, Jenks-binned",
    fill = "Value\n(Jenks bins)"
  )
```

No `scale_fill_*()` is set explicitly here - the fill comes from this
project's own `set_default_palettes()` (`R/functions.R`,
`ggplot2.discrete.fill = "viridis"`), confirmed still applying correctly
to a discrete `sf` fill exactly as it does to any other discrete
`ggplot2` scale.

A base `theme_minimal()`/`theme_void()` with the graticule stripped
(`panel.grid = element_blank()`) is the usual choropleth base - left in
for these figures since the lat/lon graticule is informative context in
a projection-comparison notebook, but for a manuscript figure the
cleaner starting point is `theme_void()`.

# Labels: `geom_sf_shadowtext()` for legibility over complex fills

A plain `geom_sf_text()`/`geom_sf_label()` label can disappear against a
busy multi-coloured background. `geom_sf_shadowtext()` (`R/functions.R`)
gives every label a white halo, the same technique
[GuangchuangYu/shadowtext](https://github.com/GuangchuangYu/shadowtext)
uses for plain `x`/`y` text - ported here to work with `sf` geometry
columns via `stat = "sf_coordinates"`, the same mechanism
`geom_sf_text()` itself uses.

```{r}
#| label: fig-shadowtext
#| fig-cap: "Country labels over a filled polygon layer, with and without a shadow"
#| fig-width: 9
#| fig-height: 5
nordics <- ne_countries(
  country = c("Norway", "Sweden", "Finland", "Denmark"),
  scale = "medium",
  returnclass = "sf"
)
nordics$centroid <- st_centroid(nordics$geometry)

p_plain <- ggplot(nordics) +
  geom_sf(aes(fill = name)) +
  geom_sf_text(aes(geometry = centroid, label = name), colour = "black", size = 4) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "geom_sf_text() - no shadow")

p_shadow <- ggplot(nordics) +
  geom_sf(aes(fill = name)) +
  geom_sf_shadowtext(
    aes(geometry = centroid, label = name),
    bg.colour = "white",
    colour = "black",
    size = 4
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "geom_sf_shadowtext() - white halo")

p_plain | p_shadow
```

Both labels sit on the same fill colours here, so the difference is
subtle at this figure's size - it stops being subtle the moment a label
lands on a genuinely dark or highly-saturated fill (a dark end of a
viridis/scico scale, in particular), which is exactly the case this
technique is for.

# Insets: `patchwork::inset_element()` vs `cowplot::draw_plot()`

Norway's NUTS2 set above included Svalbard, ~800km north of the
mainland - the standard cartographic answer is an inset, not a bounding
box that includes both and squashes the mainland. **Confirmed
`patchwork` alone can do this** - `cowplot` is not required just for
this technique, contrary to an easy assumption that `cowplot` is the
only option.

```{r}
#| label: fig-inset-patchwork
#| fig-cap: "Mainland Norway with a Svalbard inset, via patchwork::inset_element()"
#| fig-width: 6
#| fig-height: 7
mainland <- ggplot(norway) +
  geom_sf(fill = "grey85") +
  coord_sf(xlim = c(4, 32), ylim = c(57, 71.5), expand = FALSE, crs = 4326) +
  theme_minimal() +
  labs(title = "Mainland Norway")

svalbard <- ggplot(norway) +
  geom_sf(fill = "grey85") +
  coord_sf(xlim = c(9, 34), ylim = c(74, 81), expand = FALSE, crs = 4326) +
  theme_void() +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
  )

mainland + inset_element(svalbard, left = 0.55, bottom = 0.65, right = 1, top = 1)
```

```{r}
#| label: fig-inset-cowplot
#| fig-cap: "The same inset, via cowplot::draw_plot() - visually equivalent"
#| fig-width: 6
#| fig-height: 7
ggdraw(mainland) +
  draw_plot(svalbard, x = 0.55, y = 0.65, width = 0.42, height = 0.32)
```

Two things confirmed while testing this, both easy to miss:

- **`theme_void()` sets `panel.background = element_blank()`** - fully
  transparent, not just borderless. An inset built with a bare
  `theme_void()` renders with no visible fill at all against a white
  page, indistinguishable from "the inset didn't render" unless an
  explicit `panel.background = element_rect(fill = "white")` (or
  similar) is added, as above.
- **Loading both packages in one session masks
  `patchwork::align_plots()` with `cowplot`'s own version** (`cowplot`
  attached after `patchwork`, per this notebook's own setup chunk) -
  confirmed via R's own startup message, not assumed. Only relevant if
  a document actually calls `align_plots()` - `inset_element()` itself
  is unaffected - but worth knowing before reaching for that specific
  function with both packages loaded.

**Recommendation: default to `patchwork::inset_element()`**, not
`cowplot`, given this project already depends on `patchwork` for every
other plot-combining need (see [plots-mre.qmd](plots-mre.qmd)) - adding
`cowplot` only pays for itself if a specific pixel-precise placement
need comes up that `inset_element()`'s `align_to` options don't cover.

# Marine/ocean data: `mregions2` and `marmap`

General-purpose land boundaries (`rnaturalearth`/`giscoR`) have no
concept of a named sea area or an Exclusive Economic Zone - that's a
different data domain, maintained by the Flanders Marine Institute
(VLIZ) and published at marineregions.org, including the IHO's own
"Limits of Oceans and Seas" boundaries. **`mregions2` is the R-native
access point for all of it**, querying VLIZ's OGC API directly and
returning `sf`.

```{r}
#| label: fig-iho-vs-eez
#| fig-cap: "IHO Sea Areas vs Norway's actual EEZ boundary - genuinely different shapes"
#| fig-width: 8
#| fig-height: 8
iho_north <- mrp_get(
  "iho",
  cql_filter = "name IN ('Norwegian Sea', 'North Sea')"
)
eez_norway <- mrp_get("eez", cql_filter = "territory1 = 'Norway'")

ggplot() +
  geom_sf(data = iho_north, aes(fill = name), alpha = 0.4, colour = NA) +
  geom_sf(data = eez_norway, fill = NA, colour = "black", linewidth = 0.6, linetype = "dashed") +
  geom_sf(data = europe, fill = "grey90", colour = "grey40") +
  coord_sf(xlim = c(-15, 40), ylim = c(55, 82)) +
  scale_fill_brewer(palette = "Set2", name = "IHO Sea Area") +
  theme_minimal() +
  labs(
    title = "IHO Sea Areas (fill) vs Norway's EEZ (dashed outline)",
    subtitle = "A legal/economic boundary and a 1953 cartographic convention - not the same shape"
  )
```

The IHO fill and the EEZ outline visibly diverge - confirms these are
two genuinely different boundary concepts, not two sources for the same
thing. **Two gotchas found while building that figure**:

- **`mregions2::mrp_list` is a static data frame, not a function** -
  despite the naming convention of everything else in the package
  (`mrp_get()`, `mrp_view()`, ...), calling it as `mrp_list()` fails with
  `could not find function "mrp_list"`. Reference it bare (`mrp_list`)
  to browse the available layer catalogue and find the right `layer`
  code (`"iho"`, `"eez"`, ...) for `mrp_get()`.
- **IHO names don't always match the expected English spelling** -
  querying for `'%Barents%'` returns nothing; the actual name in this
  gazetteer is `"Barentsz Sea"` (the historical Dutch spelling, after
  Willem Barentsz). Use `gaz_search()` or a wildcard `LIKE` filter to
  confirm a name before assuming a `cql_filter` returned everything it
  should have.

Bathymetry is a separate concern from named-area boundaries -
**`marmap`** wraps a live NOAA/GEBCO grid fetch, and converts cleanly
into this same `sf`/`ggplot2` pipeline via `terra`/`tidyterra`:

```{r}
#| label: fig-bathymetry
#| fig-cap: "GEBCO bathymetry off the Norwegian coast, via marmap + tidyterra"
#| fig-width: 7
#| fig-height: 6
bathy <- getNOAA.bathy(
  lon1 = 0,
  lon2 = 15,
  lat1 = 58,
  lat2 = 66,
  resolution = 10,
  keep = TRUE
)
bathy_rast <- terra::rast(marmap::as.raster(bathy))

ggplot() +
  geom_spatraster(data = bathy_rast) +
  scale_fill_hypso_tint_c(
    palette = "colombia_bathy",
    guide = guide_colorbar(title = "Depth (m)")
  ) +
  theme_minimal() +
  labs(title = "Bathymetry, Norwegian coastal shelf (GEBCO via marmap)")
```

The white patches are land, not missing data - confirmed by inspecting
the raster's own elevation values there (positive, not `NA`).
`scale_fill_hypso_tint_c()`'s `_bathy`-suffixed palettes assume every
cell is a negative depth; a land cell's positive elevation falls outside
that assumed range and renders blank rather than erroring, which reads
as a data gap unless checked. Fine for a bathymetry-only figure (crop to
water before plotting, or accept blank land as context); reach for a
combined hypsometric-tint palette instead if land relief needs to show
too.

::: {.callout-note}
## Network dependency, not just a footnote

Every fetch in this section (`rnaturalearth`/`giscoR`'s first download,
`mregions2`'s OGC queries, `marmap`'s NOAA grid request) is a live
network call, made fresh on every render given this project's
`freeze: false` setting (see CLAUDE.md's "Related, separate gotcha" on
why freeze is off). `marmap::getNOAA.bathy(..., keep = TRUE)` caches its
result to a local CSV keyed by the bounding box/resolution, so repeat
renders of the *same* extent reuse it; `rnaturalearth`/`giscoR` cache
downloaded boundary files similarly. A CI runner with no prior cache, or
a network-restricted environment, will hit these servers fresh - same
category of risk already flagged for `cetz`'s package fetch in
[diagrams-mre.qmd](diagrams-mre.qmd), not independently re-verified
against a restricted CI runner here.
:::

# References
