---
# Render via tar_make(), not quarto render directly - see CLAUDE.md.
#
# To add this notebook to the nav (optional - it renders either way):
#   website:
#     navbar:
#       left:
#         - href: supplementary/notebook-template.qmd
#           text: "Supplementary: <short label>"
title: "Notebook Template"
---

# Purpose

Describe what this notebook explores and how it relates to the main
manuscript ([manuscript.qmd](../manuscript.qmd)). A notebook's layout can
mirror the manuscript (Introduction, Methods, Results) or be less
formulaic, depending on the project's needs.

```{r}
#| label: setup
#| include: false
library(targets)
library(here)
i_am("manuscript/supplementary/notebook-template.qmd")

source(here("R/functions.R"))
set_default_palettes() # viridis defaults - see colour-mre.qmd
```

{{< include _back-to-manuscript.qmd >}}

By default, console output renders as-is (plain in HTML, grey-backgrounded
in Word/PDF) - it looks bad, and you can't reference it.

```{r}
#| label: explore
# store= is required, not optional - see CLAUDE.md.
data <- tar_read(import_data, store = here("_targets"))
head(data)
```

# References