---
# ! Render this project via `targets::tar_make()` (the `render_site` target
# ! in _targets.R), not `quarto render` on this file directly — see manuscript.qmd.
#
# Files under supplementary/ are rendered as part of the site automatically
# (Quarto renders every .qmd in the project by default), but they are NOT
# added to the navbar/sidebar automatically. To surface this notebook in
# the site navigation, add an entry under `website:` in _quarto.yml, e.g.:
#
#   website:
#     navbar:
#       left:
#         - href: supplementary/notebook-template.qmd
#           text: "Supplementary: <short label>"
#
# If it's meant to stay off the public nav (e.g. a working/scratch
# notebook), leave it unlinked — it'll still render, just won't be listed.
title: "Supplementary Notebook Title"
---

# Purpose

Describe what this notebook explores and how it relates to the main
manuscript ([manuscript.qmd](../manuscript.qmd)). Depending on the needs of the
project, the layout of a notebook may be a miniature version of the main manuscript (Introduction, Methods, Results, etc.) or something less formulaic.

```{r}
#| label: setup
#| include: false

# It's generally wise to set include: false wherever you load your packages, as the function returns a lot of session-specific information which is rarely relevant to documents
library(targets)
library(here)

# i_am()'s argument is THIS file's own path relative to the project root -
# not _targets.R's path. Anchors here() to the root regardless of where this
# notebook gets knit from.
i_am("manuscript/supplementary/notebook-template.qmd")
```

{{< include _back-to-manuscript.qmd >}}

By default code chunk output to the console is rendered as is (straight onto a white background in HTML, although it gets a grey background in Word and PDF). It looks pretty bad. You also can't reference it.

```{r}
#| label: explore
# store= is required here, not optional: tar_read()'s default store
# lookup is cwd-relative, not here()-anchored, and cwd during chunk
# execution isn't reliable across every render path (confirmed: it's
# manuscript/ under a direct `quarto render`, but the repo root under
# this project's actual tarchetypes::tar_quarto() render - see CLAUDE.md).
# here("_targets") works either way since here() resolves via upward
# sentinel search, not cwd.
data <- tar_read(simulate_data, store = here("_targets"))
head(data)
```

# References