---
# ! Render via `targets::tar_make()` (the `render_manuscript` target in
# ! _targets.R), not `quarto render` directly - see manuscript.qmd.
#
# Files under supplementary/ render as part of the site automatically
# (Quarto renders every .qmd by default) but aren't added to the
# navbar/sidebar automatically. To surface this notebook in the nav, add
# an entry under `website:` in _quarto.yml:
#
#   website:
#     navbar:
#       left:
#         - href: supplementary/notebook-template.qmd
#           text: "Supplementary: <short label>"
#
# Leave it unlinked to keep it off the public nav (e.g. a scratch
# notebook) - it still renders, just isn't listed.
title: "Supplementary Notebook Title"
---

# Purpose

Describe what this notebook explores and how it relates to the main
manuscript ([manuscript.qmd](../manuscript.qmd)). A notebook's layout can
mirror the manuscript (Introduction, Methods, Results) or be less
formulaic, depending on the project's needs.

```{r}
#| label: setup
#| include: false

# Set include: false wherever you load packages - library() output is
# session-specific noise, rarely relevant to the document.
library(targets)
library(here)

# i_am()'s argument is THIS file's own path relative to the project root,
# not _targets.R's. Anchors here() regardless of where this notebook is knit from.
i_am("manuscript/supplementary/notebook-template.qmd")
```

{{< include _back-to-manuscript.qmd >}}

By default, console output renders as-is (plain in HTML, grey-backgrounded
in Word/PDF) - it looks bad, and you can't reference it.

```{r}
#| label: explore
# store= is required, not optional: tar_read()'s default store lookup is
# cwd-relative, and cwd during chunk execution isn't reliable across
# render paths (confirmed: manuscript/ under a direct `quarto render`, but
# the repo root under this project's actual tar_quarto() render - see
# CLAUDE.md). here("_targets") works either way since here() resolves via
# upward sentinel search, not cwd.
data <- tar_read(import_data, store = here("_targets"))
head(data)
```

# References