---
# Render via `targets::tar_make()` (the `render_manuscript` target in
# _targets.R), not `quarto render` directly - see manuscript.qmd.
#
# Minimal reproducible example (MRE) for ggplot2 graphics in this
# template: a session-wide theme, ggtext markdown text (including a real,
# confirmed bug), and combining plots with patchwork - including a table
# grob alongside a plot grob.
title: "ggplot2: session themes, markdown text, and patchwork"
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(ggtext)
library(patchwork)
library(flextable)
library(here)
here::i_am("manuscript/supplementary/plots-mre.qmd")

# Sets the default theme for every ggplot built in the REST OF THIS
# DOCUMENT's render session - no need to add + theme_something() to each
# plot individually. Scoped to this notebook's own knitr session only
# (see CLAUDE.md's "Pipeline gotchas" for why each .qmd/partial is its own
# render context) - it does not affect other notebooks or manuscript.qmd.
theme_set(theme_minimal(base_size = 12))
```

{{< include _back-to-manuscript.qmd >}}

::: {.callout-tip}
# The Short Version
- `theme_set()` in the setup chunk applies to every later chunk in the
  same render - no need to add `+ theme_something()` to each plot.
- `ggtext::element_markdown()` lets theme text elements (titles,
  subtitles, axis text, captions) render limited Markdown - `**bold**`,
  `*italic*`, and more.
- **Confirmed real bug, still present in ggplot2 4.0.3**: adding another
  *complete* theme (e.g. `+ theme_bw()`) after `theme_set()`'d an
  `element_markdown()` override silently wipes it back to plain text -
  not a rendering glitch, a real class change
  (`element_markdown` → `element_text`), confirmed by inspecting
  `plot$theme$plot.title`'s class directly, not just by eye. Root cause:
  ggplot2 *replaces*, not merges, when one complete theme is added on top
  of another.
- Fix: reapply the `element_markdown()` override *after* any later
  complete theme - every time, since order is what matters, not where the
  original override came from.
- `patchwork` combines multiple plots (and other grid grobs) with
  `+`/`/`/`()` and `plot_layout()`/`plot_annotation()`.
- A table can share a `patchwork` layout with a plot two ways:
  `patchwork::wrap_table()` for a plain data frame, or
  `flextable::gen_grob()` + `patchwork::wrap_elements()` for a fully
  styled `flextable` - ties into this project's own table conventions,
  see [tables-mre.qmd](tables-mre.qmd).
:::

# Why this notebook exists

[tables-mre.qmd](tables-mre.qmd) and [images-mre.qmd](images-mre.qmd) are
both about *format support* - does the same code produce equivalent
output across html/docx/typst. This notebook is a different kind of
concern: session-wide styling conventions, and a genuine ggplot2 gotcha
that's easy to hit and easy to misdiagnose (it looks like `ggtext` isn't
working at all, when the actual cause is theme ordering elsewhere in the
document). Findings below are from direct testing in this template, not
assumed from documentation.

# Basic ggplot graphs

Nothing template-specific here - just confirming the session theme (set
in the setup chunk above) applies with no `+ theme_something()` needed on
any individual plot:

```{r}
#| label: fig-scatter
#| fig-cap: "A basic scatter plot, using the session-wide theme"
ggplot(mtcars, aes(wt, mpg, colour = factor(cyl))) +
  geom_point(size = 2) +
  labs(x = "Weight (1000 lbs)", y = "Miles per gallon", colour = "Cylinders")
```

```{r}
#| label: fig-bar
#| fig-cap: "A basic bar chart, same session-wide theme, no extra code"
ggplot(mtcars, aes(factor(cyl))) +
  geom_bar(fill = "steelblue") +
  labs(x = "Cylinders", y = "Count")
```

# A session-wide theme

`theme_set(theme_minimal(base_size = 12))` in the setup chunk (`include:
false`, so it doesn't print) is what both plots above picked up
automatically. This is `ggplot2`'s own intended mechanism for this -
`theme_set()` changes the *default* theme returned by `theme_get()`,
which every subsequent plot in the same R session is built against unless
it adds its own theme explicitly. Confirmed scoped correctly: this
setting does not leak into other notebooks or `manuscript.qmd`, since
each is rendered in its own knitr session (see CLAUDE.md's "Pipeline
gotchas").

# Markdown text with ggtext

`ggtext::element_markdown()` swapped in for a theme text element renders
limited Markdown instead of a literal string:

```{r}
#| label: fig-markdown-title
#| fig-cap: "A title with **bold** and *italic* markdown, rendered correctly"
p_markdown <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(title = "**Weight** vs. *fuel economy*") +
  theme(plot.title = element_markdown())
p_markdown
```

The title above renders "Weight" bold and "fuel economy" italic, not the
literal asterisks - confirmed by inspecting the rendered output directly,
not just trusting the code looks right.

## The bug: a later complete theme silently drops the markdown override

`theme_bw()`, `theme_minimal()`, and friends are **complete** themes -
`attr(theme_bw(), "complete")` is `TRUE`. A bare `theme(plot.title =
element_markdown())` is **not** complete
(`attr(theme(plot.title = element_markdown()), "complete")` is `FALSE`) -
it's a partial modification, meant to be merged onto whatever theme is
already active. Confirmed directly: adding one complete theme on top of
another **replaces** the existing theme wholesale rather than merging -
this is what breaks the pattern above the moment a second complete theme
enters the picture, and it's easy to hit by accident (reaching for `+
theme_bw()` on one specific plot for contrast, say) without realising it
silently undoes an earlier `element_markdown()` override set via
`theme_set()` further up the document.

```{r}
#| label: fig-markdown-bug
#| fig-cap: "Same markdown title, after + theme_bw() - reverts to literal asterisks"
#| fig-width: 8
p_broken <- p_markdown + theme_bw()
p_broken
```

```{r}
#| label: check-theme-classes
cat(
  "p_markdown's plot.title element class:",
  class(p_markdown$theme$plot.title)[1],
  "\n"
)
cat(
  "p_broken's plot.title element class:  ",
  class(p_broken$theme$plot.title)[1],
  "\n"
)
```

Confirms it's a real class change, not a rendering fluke: `plot.title`
goes from `element_markdown` to a plain `element_text` the moment
`theme_bw()` (complete) is added after the fact - `theme_bw()` supplies
its *own* `plot.title` (an ordinary `element_text`), and since adding one
complete theme replaces rather than merges, that plain element wins,
regardless of what was active before.

**Fix: reapply the markdown override *after* any later complete theme.**
Order is what matters here, not where the original override came from:

```{r}
#| label: fig-markdown-fixed
#| fig-cap: "theme_bw() + the markdown override reapplied afterward - fixed"
#| fig-width: 8
p_fixed <- p_markdown + theme_bw() + theme(plot.title = element_markdown())
p_fixed
```

There's no way to make the override "stick" through an arbitrary later
complete theme automatically - reapplying it after every complete theme
added downstream is the only fix, since the replacement happens at
`+`-time on whatever complete theme arrives last, not at `theme_set()`
time.

# Combining plots with patchwork

`patchwork` combines separate ggplot objects with `+` (wrap into an
implicit grid), `/` (stack vertically), and `plot_layout()`/
`plot_annotation()` for finer control:

```{r}
#| label: fig-patchwork
#| fig-cap: "Three plots combined with patchwork - side by side, then stacked"
#| fig-width: 8
#| fig-height: 6
p_a <- ggplot(mtcars, aes(wt, mpg)) + geom_point() + labs(title = "A")
p_b <- ggplot(mtcars, aes(factor(cyl))) + geom_bar() + labs(title = "B")
p_c <- ggplot(mtcars, aes(mpg)) + geom_histogram(bins = 10) + labs(title = "C")

(p_a | p_b) /
  p_c +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "patchwork: (A | B) / C",
    tag_levels = "1"
  )
```

# Combining a plot grob and a table grob

Two ways to put a table alongside a plot in the same `patchwork` layout.
The quick way, for a plain data frame - `patchwork::wrap_table()`:

```{r}
#| label: fig-wrap-table
#| fig-cap: "A plot beside a plain data frame, via patchwork::wrap_table()"
#| fig-width: 8
p_a + wrap_table(head(mtcars[, c("mpg", "wt", "cyl")], 5))
```

The more capable way, for a fully-styled `flextable` - convert it to a
grob with `flextable::gen_grob()`, then hand that grob to
`patchwork::wrap_elements()`. This is the version worth using when the
table needs the same styling conventions as the rest of the manuscript
(see [tables-mre.qmd](tables-mre.qmd) for why `flextable` is this
project's default table package):

```{r}
#| label: fig-flextable-grob
#| fig-cap: "A plot beside a styled flextable, via gen_grob() + wrap_elements()"
#| fig-width: 8
ft <- flextable(head(mtcars[, c("mpg", "wt", "cyl")], 5)) |>
  theme_vanilla() |>
  flextable::bold(part = "header") |>
  autofit()
tbl_grob <- flextable::gen_grob(ft, fit = "fixed", scaling = "min")

p_a + wrap_elements(full = tbl_grob)
```

Both are ordinary `grid` grobs by the time `patchwork` draws them, so
this isn't format-specific the way file-based images are (see
[images-mre.qmd](images-mre.qmd)) - knitr rasterises the whole combined
figure through the standard R graphics device regardless of html/docx/
typst, confirmed by rendering this notebook in all three and checking
each output directly rather than assuming grid composition "just works"
everywhere.

# References
