---
# Render this project via `targets::tar_make()` (the `render_site` target in
# _targets.R), not `quarto render` on this file directly — see index.qmd.
#
# This is a minimal reproducible example (MRE) for comparing R's main
# table packages side by side across HTML, docx, and typst (PDF) output.
# Render all three formats and open them next to each other when
# deciding which package to standardise on.
#
# To surface this notebook in the site navigation, see the comment in
# supplementary/notebook-template.qmd.
title: "Tables in Quarto: Packages, formats, and bugs"
---

```{r}
#| label: setup
#| include: false
library(knitr)
library(kableExtra)
library(gt)
library(flextable)
library(huxtable)
library(DT)

# Tested directly in this template (not assumed from docs): which output
# a package actually degrades to depends on the specific target format,
# not just "HTML vs. not-HTML" — gt works for typst but not docx, for
# example. Gate chunks on the real target rather than guessing.
pandoc_to <- knitr::opts_knit$get("rmarkdown.pandoc.to")
is_html <- identical(pandoc_to, "html")
is_typst <- identical(pandoc_to, "typst")
is_docx <- identical(pandoc_to, "docx")

quarto_env <- Sys.getenv("QUARTO_EXECUTE_INFO")

set.seed(1)
dat <- data.frame(
  Group = rep(c("Control", "Treatment"), each = 3),
  Metric = rep(c("Alpha", "Beta", "Gamma"), 2),
  Value = round(rnorm(6, mean = 10, sd = 2), 2),
  p = round(runif(6, 0, 0.1), 3)
)

# A "big stupid table" — enough rows and long enough text to actually
# stress column widths, wrapping and pagination, instead of the 6-row toy
# example above which hides all of that.
words <- c(
  "alpha",
  "beta",
  "gamma",
  "delta",
  "epsilon",
  "zeta",
  "eta",
  "theta",
  "iota",
  "kappa",
  "lambda",
  "mu",
  "nu",
  "xi",
  "omicron",
  "pi",
  "rho",
  "sigma",
  "tau",
  "upsilon"
)
make_note <- function() {
  n_sentences <- sample(1:3, 1)
  paste(
    replicate(n_sentences, {
      n_words <- sample(8:20, 1)
      s <- paste(sample(words, n_words, replace = TRUE), collapse = " ")
      paste0(toupper(substring(s, 1, 1)), substring(s, 2), ".")
    }),
    collapse = " "
  )
}
big <- data.frame(
  ID = seq_len(30),
  Group = sample(
    c("Control", "Treatment A", "Treatment B"),
    30,
    replace = TRUE
  ),
  Score = round(rnorm(30, 50, 10), 1),
  Notes = vapply(seq_len(30), \(i) make_note(), character(1))
)

# A bigger dataset again, for the DT pagination demo specifically.
huge <- data.frame(
  ID = seq_len(250),
  Group = sample(
    c("Control", "Treatment A", "Treatment B"),
    250,
    replace = TRUE
  ),
  Score = round(rnorm(250, 50, 10), 1)
)
```

Quarto render env is:

```{r}
quarto_env
```


::: {.callout-tip}
# The Short Version
- Use `flextable` for reasonably good HTML/docx/typst tables
- Use `dt` where interactive data exploration is needed
- Tables of more than one page break Typst rendering across packages
- To get nice tables in docx, specify column widths manually
- Never assume formatting that works with a small table will work on a big ugly one
:::

# Why this notebook exists

Tables are a pain — in Markdown, in HTML, in Word, in PDF, and
especially across all three from one source document. The same table
code does **not** produce equivalent output across formats; several
packages fail outright on some targets rather than degrading gracefully.
Notes below are from direct testing in this template, not general
folklore — commands were actually run against `--to html`, `--to docx`,
and `--to typst` while writing this.

# The Riders & Runners

## Plain markdown tables

No R involved — pandoc's own table syntax, useful for small
hand-written tables that don't need computed values. Two forms:

**Pipe table:**

| Group     | Metric | Value |
|-----------|--------|------:|
| Control   | Alpha  |  9.83 |
| Control   | Beta   | 11.20 |
| Treatment | Alpha  | 10.55 |

**Grid table** (supports multi-line/multi-paragraph cells, pipe tables
don't):

+------------+-------------------------------------------------+
| Group      | Notes                                           |
+============+=================================================+
| Control    | Baseline arm.                                   |
|            |                                                 |
|            | No intervention applied.                        |
+------------+-------------------------------------------------+
| Treatment  | Received the intervention starting week 2.      |
+------------+-------------------------------------------------+

- ✅ Both render natively in html/docx/typst with no package involved —
- same pandoc table writer as `kable()`'s output, so docx `Table` style caveat above still applies to these too.

Grid tables are genuinely fragile to hand-write: the `+`/`|` characters
must line up *exactly* by column position across every row. The first
draft of the table above had one column one space too narrow in the
header row — pandoc didn't error, it silently dropped the entire
"Baseline arm." paragraph from the rendered output, with no warning.
`knitr::kable(x, format = "grid")` looks like the obvious way to
generate one instead of hand-aligning — it didn't work in the version
tested here (`could not find function "kable_grid"`). Safest options:
hand-align very carefully and check the rendered output, or just use a
pipe table if multi-paragraph cells aren't actually needed.

## `knitr::kable()`

- Plain markdown/grid table, no package-specific styling.
- ✅ Works everywhere (html/docx/typst) because it's just pandoc's own
  table syntax — but with zero control over appearance. In docx it
  renders using whatever the `Table` style resolves to in the reference
  doc (see the README's `db-space-line-n.docx` notes).

```{r}
#| label: tbl-kable
#| tbl-cap: "kable"
kable(dat)
```

## `kableExtra`

- ✅ Pretty HTML via Bootstrap classes.
- 🛑 **Hard-errors** the whole render for both docx and typst — not a
  silent no-op. Confirmed with `kable_styling(bootstrap_options = ...)`
  alone, no other package involved: `Functions that produce HTML output
  found in document targeting docx output`. Same failure for typst.
- Bootstrap styling has no docx/typst equivalent to fall back to, so
  there's nothing sensible for it to degrade into anyway — gate it to
  HTML rather than expecting a fallback.

```{r}
#| label: tbl-kableextra
#| tbl-cap: "kableExtra"
#| eval: !expr is_html
kable(dat) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) |>
  row_spec(0, bold = TRUE)
```

::: {.content-visible unless-format="html"}
*`kableExtra` output omitted here — HTML-only, see the note above.*
:::

## `gt`

- ✅ Pretty HTML with conditional colouring.
- 🛑 Hard-errors for **docx** (same "HTML output found" error as
  kableExtra), *unless* the `webshot2` package is installed, in which
  case it falls back to a static raster screenshot — not an editable
  Word table.
- ✅ Renders **natively to typst** — confirmed, no error, real vector
  table (not a screenshot). This differs from docx and is easy to
  assume incorrectly if you've only tested one non-HTML format.

```{r}
#| label: tbl-gt
#| tbl-cap: "gt"
#| eval: !expr (is_html || is_typst)
gt(dat) |>
  tab_header(title = "gt table") |>
  fmt_number(columns = c(Value, p), decimals = 2) |>
  data_color(columns = Value, palette = "viridis")
```

::: {.content-visible when-format="docx"}
*`gt` output omitted here — see the note above (works for html/typst,
not docx).*
:::

## `flextable`

- ✅✅✅ The most reliable cross-format option. Targets the Word OOXML table
  model directly (via `officer`), and also renders cleanly to HTML and
  typst — confirmed working in all three with no extra configuration.
- No Lua filter needed in the YAML frontmatter for any of the three
  formats tested here — Quarto's built-in `knit_print.flextable` /
  print-method integration handles it automatically. (If you've seen
  advice elsewhere about registering a flextable filter, it isn't
  necessary for html/docx/typst as of the Quarto version this was
  tested with.)

```{r}
#| label: tbl-flextable
#| tbl-cap: "flextable"
# huxtable and flextable both export bold()/width()/etc. with different
# signatures — loading both means whichever loaded last wins by default.
# Namespace explicitly rather than relying on load order (bit us during
# testing: an unqualified width() call below picked up huxtable's version
# and errored with "unused arguments").
flextable(dat) |>
  theme_vanilla() |>
  flextable::bold(part = "header") |>
  autofit()
```

## `huxtable`

- ✅✅✅ Also genuinely cross-format (html/docx/typst) from one spec, smaller
  styling vocabulary than flextable.
- Quirk found in typst specifically: the table renders left-shifted
  relative to `gt`/`flextable`'s centred output on the same page — check
  it visually rather than assuming it matches your other tables in PDF.

```{r}
#| label: tbl-huxtable
#| tbl-cap: "huxtable"
hux <- as_huxtable(dat)
bold(hux)[1, ] <- TRUE
bottom_border(hux)[1, ] <- 1
hux
```

## `DT` (interactive, HTML only)

- Not a static table — a paginated, sortable, filterable JS widget
  (DataTables). This is the one to reach for when a table is too big to
  dump on the page at once, rather than trying to force `kable`/`gt` to
  cope with hundreds of rows.
- HTML-only by nature (it's a JS widget) — same "HTML output" failure
  as kableExtra/gt if you don't gate it for docx/typst.

```{r}
#| label: tbl-dt
#| tbl-cap: "DT"
#| eval: !expr is_html
datatable(huge, options = list(pageLength = 10), filter = "top")
```

::: {.content-visible unless-format="html"}
*`DT` output omitted here — it's an HTML/JS widget with no static
docx/typst equivalent. `huge` has 250 rows; this is exactly the case
where pagination beats a giant static table.*
:::

## Stress test: a big, ugly table

30 rows, a `Notes` column with 1–3 sentences of variable-length text.
Small toy tables hide exactly the problems that show up here — this is
what actually surfaced the worst finding in this whole notebook, below.

::: {.callout-warning}
## Found while stress-testing: typst mangles tables that span a page break

Render this notebook's PDF and look at the stress-test tables (`kable`,
`flextable`, and `gt` — all three, not just one package): wherever the
30-row table would break across a page boundary, the text becomes
unreadable, interleaved character-soup for the rows around the break.
Confirmed with raw text extraction (`pdftotext`), not just a PDF-viewer
rendering glitch. It does not happen in HTML or docx — this is
typst-specific, and it does not error or warn; the render "succeeds"
with corrupted output.

Reproduced with Quarto 1.9.36. Root cause looks like Quarto's
markdown/pandoc-to-typst table writer not handling automatic
page-breaking correctly once wrapped multi-line cell content is
involved — small tables (the toy `dat` examples above) never hit this
because they fit on one page.

**If you're targeting typst/PDF: keep tables short enough to fit on one
page**, or split a long table into page-sized chunks, or exclude it from
the typst render entirely (gate with `eval: !expr !is_typst`, same
pattern used elsewhere in this notebook) and point PDF readers at the
HTML/docx version instead. Don't assume a clean `quarto render` with no
errors means the PDF is actually correct — check it.
:::

**`kable`** — no width control at all. In docx this comes out as a
plain grid table sized however Word feels like sizing it; long `Notes`
text is not reliably wrapped in a readable way.

```{r}
#| label: tbl-big-kable
#| tbl-cap: "Stress test: kable"
kable(big)
```

**`flextable`** — same package as above, but here explicit column
widths are set instead of `autofit()`. `autofit()` sizes every column to
fit its content on one line, which for a long-text column means the
table blows past the page width; fixed `width()` + flextable's automatic
text wrapping (row height grows instead) is the fix.

```{r}
#| label: tbl-big-flextable
#| tbl-cap: "Stress test: flextable with explicit column widths"
flextable(big) |>
  theme_vanilla() |>
  flextable::bold(part = "header") |>
  flextable::width(j = c("ID", "Group", "Score"), width = 0.8) |>
  flextable::width(j = "Notes", width = 3.5) |>
  flextable::fontsize(size = 9, part = "all")
```

**`gt`** (html/typst only) — no column-width fix applied here
deliberately, so you can see the same overflow risk `kable` has when a
text column isn't constrained.

```{r}
#| label: tbl-big-gt
#| tbl-cap: "Stress test: gt"
#| eval: !expr (is_html || is_typst)
gt(big)
```

::: {.content-visible when-format="docx"}
*`gt` output omitted here — see the `gt` section above.*
:::

**`DT`** (html only) — this is the actual answer to "the table is too
big": don't fight a static format, paginate it.

```{r}
#| label: tbl-big-dt
#| tbl-cap: "Stress test: DT"
#| eval: !expr is_html
datatable(big, options = list(pageLength = 5))
```

::: {.content-visible unless-format="html"}
*`DT` output omitted here — HTML-only, see the note above.*
:::

**Takeaway:** docx handles a long table running across pages fine
(native Word behaviour). typst does not — see the callout above, this
is the real finding from stress-testing, not the column-width stuff.
For HTML, don't use a static table package for anything much bigger
than fits on one screen — use `DT`.

