---
# ! Render via `targets::tar_make()` (the `render_manuscript` target in
# ! _targets.R), not `quarto render` directly - see manuscript.qmd.
title: "Testing R/functions.R without a package structure"
---

```{r}
#| label: setup
#| include: false
library(here)
here::i_am("manuscript/supplementary/testing-mre.qmd")
```

{{< include _back-to-manuscript.qmd >}}

::: {.callout-tip}
# The Short Version
- Tests live in `tests/testthat/`, run with `testthat::test_dir("tests/testthat")`
  - there's no `DESCRIPTION`/`NAMESPACE` here, so the usual package-testing
  entry points (`devtools::test()`, `R CMD check`) don't apply.
- `tests/testthat/helper.R` sources `R/functions.R` once for every test
  file - the stand-in for `library(pkg)`, which a package's own tests get
  for free.
- Coverage is deliberately barebones: `detect_render_format()` (the one
  function with real, previously-buggy branch logic - see its own roxygen
  comment) and a smoke test per export helper that just checks a file
  actually gets written.
- `.github/workflows/render.yml` runs the suite before the pipeline build,
  so a broken helper function fails as a named test rather than a
  confusing render error - see [ci-pipeline.qmd](ci-pipeline.qmd).
:::

# Why this notebook exists

`R/functions.R` has real branch logic worth protecting against
regressions - `detect_render_format()`'s own roxygen comment documents a
concrete bug a previous version had (silently broken whenever
`QUARTO_EXECUTE_INFO` was unset). That kind of bug is exactly what a test
catches immediately and a manual render might not, since the broken branch
only triggers in specific circumstances (a standalone knit, or
`tar_quarto()`'s dependency-scan pass - see CLAUDE.md's "Pipeline
gotchas"). This notebook exists because "why is there a `tests/` folder in
a manuscript template, and why does it look different from a package's
tests" isn't self-explanatory from the files alone.

# Why not a normal package test setup

The standard `usethis::use_testthat()` workflow assumes a package: a
`DESCRIPTION` with `Suggests: testthat`, a `NAMESPACE`, and `R CMD check`
or `devtools::test()` doing the loading for you via `pkgload`. This
template deliberately isn't a package (see CLAUDE.md's "Core architectural
decision" for the equivalent reasoning about the Quarto project type) -
adding full package scaffolding just to get test infrastructure would be a
lot of ceremony for five tests. Instead:

- `tests/testthat/helper.R` is testthat's own convention for a file
  auto-loaded before every test file in the directory - normally used for
  shared fixtures in a package's tests, repurposed here to do the one
  thing `devtools::load_all()` would otherwise handle: source
  `R/functions.R` so its functions exist in the test session.
- Tests are run directly with `testthat::test_dir("tests/testthat")`,
  which only needs a directory of `test-*.R` files - it doesn't care
  whether a package wraps them.

```{r}
#| label: run-tests
#| eval: false
testthat::test_dir(here("tests/testthat"))
```

(Not run as part of the manuscript render - `eval: false` above - since a
test failure should stop CI before the render even starts, per
[ci-pipeline.qmd](ci-pipeline.qmd), not surface as a broken chunk in the
rendered notebook.)

# What's actually tested

`tests/testthat/test-detect-render-format.R` - three cases:

- `QUARTO_EXECUTE_INFO` unset → every `is_*` flag `FALSE`, `ext` is `NA`,
  not an error or a missing field. This is the exact contract the
  function's roxygen comment promises, and the exact thing the previous
  hand-rolled version got wrong.
- A JSON file with `target-format: "docx"` → `is_docx` `TRUE`, the other
  two flags `FALSE`, `ext = "docx"`.
- A JSON file with `target-format: "typst"` → `ext = "pdf"`, not
  `"typst"` - typst compiles to a PDF on disk (see
  [rendering-pipeline.qmd](rendering-pipeline.qmd)), and this is the one
  format/extension mismatch in the whole function worth pinning down
  explicitly rather than trusting it stays correct by accident.

`tests/testthat/test-exports.R` - one smoke test each for
`export_table_docx()` and `export_figure_tiff()`: build a trivial
flextable/ggplot object, export it to a `withr::local_tempdir()`, check
the expected file actually exists at the expected path. Deliberately not
testing docx/TIFF internals - that's what
[tables-mre.qmd](tables-mre.qmd)/[images-mre.qmd](images-mre.qmd) already
do by rendering real output and inspecting it by hand; these tests only
catch "this function no longer writes a file where it claims to."

**Not tested:** `flextable_use_format_font()`. Checking it picked the
right font would mean introspecting a flextable object's internal
`fp_text` structure per format, which is more machinery than "barebones"
was meant to cover - if this function grows more format-specific branches,
it's the next candidate for a real test.

# References
