---
# Render this project via `targets::tar_make()` (the `site` target in
# _targets.R), not `quarto render` on this file directly — see index.qmd.
#
# This is a minimal reproducible example (MRE) for comparing R's main
# table packages side by side across HTML and Word (docx) output. Render
# both formats and open output/supplementary/tables-mre.html and .docx
# next to each other when deciding which package to standardise on.
#
# To surface this notebook in the site navigation, see the comment in
# supplementary/notebook-template.qmd.
title: "Tables MRE: kable / kableExtra / gt / flextable / huxtable"
---

```{r}
#| label: setup
#| include: false
library(knitr)
library(kableExtra)
library(gt)
library(flextable)
library(huxtable)

set.seed(1)
dat <- data.frame(
  Group = rep(c("Control", "Treatment"), each = 3),
  Metric = rep(c("Alpha", "Beta", "Gamma"), 2),
  Value = round(rnorm(6, mean = 10, sd = 2), 2),
  p = round(runif(6, 0, 0.1), 3)
)
```

# Why this notebook exists

NB: Word output is still corrupted as of this version.

Tables are a pain. In general, in Markdown, in HTML, in Word, and especially in pandoc-generated .docx files.

The same table code does **not** produce equivalent output in HTML vs.
docx — each package has a different HTML-vs-Word story. Notes below are
from direct testing in this template.
# `knitr::kable()`

`knitr::kable()`

- returns a markdown pipe table
- no special formatting
- supported by HTML (pretty) and Word (unstyled (?), not great)

```{r}
#| label: tbl-kable
#| tbl-cap: "kable"
kable(dat)
```

# `kableExtra`

- returns pretty HTML tables
- supports Bootstrap
- 🛑 crashes the docx render entirely

::: {.content-visible when-format="docx"}

```{r}
#| label: tbl-kableextra
#| tbl-cap: "kableExtra"
#| eval: !expr knitr::is_html_output()
kable(dat) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) |>
  row_spec(0, bold = TRUE)
```

*`kableExtra` output omitted here — see the note above.*
:::

# `gt`

- pretty HTML with colouring
- 🛑 rendering to docx crashes the renderer (unless `webshot2` is installed, then returns static image)

::: {.content-visible when-format="docx"}

```{r}
#| label: tbl-gt
#| tbl-cap: "gt"
#| eval: !expr knitr::is_html_output()
gt(dat) |>
  tab_header(title = "gt table") |>
  fmt_number(columns = c(Value, p), decimals = 2) |>
  data_color(columns = Value, palette = "viridis")
```

*`gt` output omitted here — see the note above.*
:::

# `flextable`

- generally pretty HTML (doesn't respect Quarto/CSS styles?) and .docx output
- requires flextable Lua filter to be set in frontmatter (although it isn't here???)
- targets the Word OOXML table model directly (via `officer`)

```{r}
#| label: tbl-flextable
#| tbl-cap: "flextable"
# huxtable::bold() and flextable::bold() collide once both packages are
# loaded — namespace explicitly rather than relying on load order.
flextable(dat) |>
  theme_vanilla() |>
  flextable::bold(part = "header") |> # namespace collision
  autofit()
```

# `huxtable`

- also pretty HTML (but not well spaced?) and .docx output
- I've never used this before, but it looks fine?

```{r}
#| label: tbl-huxtable
#| tbl-cap: "huxtable"
hux <- as_huxtable(dat)
bold(hux)[1, ] <- TRUE
bottom_border(hux)[1, ] <- 1
hux
```

# Recommendation
